### 2015-10-10 paper/main.tex
Working-ish prototype.

\documentclass{sigplanconf}

\usepackage{amsmath}
\usepackage{tabularx}
\usepackage{listings}
\lstset{
  language=caml,
  basicstyle=\ttfamily,
}
\def\li{\lstinline}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
\copyrightyear{20yy}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish,
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers,
                                  % short abstracts)

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Function pearl: the proof search monad}
\subtitle{Subtitle Text, if any}

\authorinfo{Jonathan Protzenko}
           {Microsoft Research}
           {protz@microsoft.com}

\maketitle

\begin{abstract}
  We present the proof search monad: the combination of the option monad, the
  backtracking monad, and the sequencing monad. The proof search monad allows
  writing a proof search engine that looks like the actual derivation rules,
  while generating a proof tree at the same time.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore,
% you may leave them out
\terms
term1, term2

\keywords
keyword1, keyword2

\section{Conjunctions of equalities}

Just the regular option monad. Lazyness is already built-in (the second
computation is suspended).

In this first version, we treat the simple problem of conjunction of
equalities\dots written in an elegant style with a persistent union-find,
functional, the option monad, because we may fail.

Lazyness already built-in, evaluation stops right after the first failing
premise has been hit.

Goal: build a derivation tree at the same time. Requires: recording the
premises, in order to build an application of the rule.

\bibliographystyle{abbrvnat}
\bibliography{local}

\end{document}

### 2015-10-12 paper/main.tex
Snapshot + paper.

@@
-\documentclass{sigplanconf}
+\documentclass{easychair}
 
 \usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{stmaryrd}
+\usepackage{amstext}
+\usepackage{amsthm}
+\usepackage{hyperref}
+
 \usepackage{tabularx}
 \usepackage{listings}
+  \def\li{\lstinline}
   \lstset{
     language=caml,
+    flexiblecolumns=false,
+    showstringspaces=false,
     basicstyle=\ttfamily,
+    framesep=5pt,
+    numberstyle=\tiny,
+    numbersep=5pt,
+    escapeinside={(*}{*)},
+    morekeywords={above,abstract,adopts,alias,and,as,assert,begin,below,builtin,consumes,data,do,downto,duplicable,dynamic,else,empty,end,exclusive,explain,fact,fail,flex,for,from,fun,give,if,in,let,match,mutable,open,pack,perm,preserving,rec,take,taking,then,to,type,unknown,val,value,while,with,witness},
+    deletekeywords={value},
   }
-\def\li{\lstinline}
+\usepackage{minted}
+  \newminted[ocaml]{ocaml}{mathescape}
+\usepackage{mathpartir}
+  % Improvements to Didier's mathpartir package: make all rule names and
+  % references use textsc, and also make references to rules done using the \Rule
+  % command clickable in a PDF.
+  \let\TirName\textsc
+  \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
+  \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
+  \let\DefRule\RefTirName
+  \let\Rule\DefTirName
+
+\newcommand{\fref}[1]{Figure~\ref{fig:#1}}
+\newcommand{\R}{\ensuremath{\mathcal{R}}} % Rigid
+\newcommand{\f}[1]{\ensuremath{\mathbf{#1}}} % Bold
+\newcommand{\F}{\ensuremath{\mathcal{F}}} % Flexible
+\newcommand{\V}{\ensuremath{\mathcal{V}}} % The prefix
 
@@
 
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
+\title{Functional Pearl: the Proof Search Monad}
+\titlerunning{The Proof Search Monad}
 
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
-\copyrightyear{20yy}
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
-\doi{nnnnnnn.nnnnnnn}
-
-% Uncomment one of the following two, if you are not going for the
-% traditional copyright transfer agreement.
-
-%\exclusivelicense                % ACM gets exclusive license to publish,
-                                  % you retain copyright
-
-%\permissiontopublish             % ACM gets nonexclusive license to publish
-                                  % (paid open-access papers,
-                                  % short abstracts)
-
-\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{short description of paper}   % 'preprint' option specified.
-
-\title{Function pearl: the proof search monad}
-\subtitle{Subtitle Text, if any}
-
-\authorinfo{Jonathan Protzenko}
-           {Microsoft Research}
-           {protz@microsoft.com}
+\author{Jonathan Protzenko}
+\authorrunning{J. Protzenko}
+\institute{
+  Microsoft Research\\
+  \email{protz@microsoft.com}}
 
@@
 \begin{abstract}
-  We present the proof search monad: the combination of the option monad, the
-  backtracking monad, and the sequencing monad. The proof search monad allows
-  writing a proof search engine that looks like the actual derivation rules,
-  while generating a proof tree at the same time.
+  We present the proof search monad, a set of combinators that allows one to
+  write a proof search engine in a style that resembles closely the inference
+  rules. The user calls functions such as \li+premise+, \li+prove+ or
+  \li+choice+; the library then takes care of generating a derivation tree,
+  which serves either as a witness (success) or as an error message (failure).
+  Proof search engines written in this style enjoy: first, a one-to-one
+  correspondence between the implementation and the theoretical rules, which
+  makes manual inspection easier; second, proof witnesses ``for free'', which
+  makes a verified-validation approach easier too.
 \end{abstract}
 
-\category{CR-number}{subcategory}{third-level}
+\section{A minimal problem}
 
-% general terms are not compulsory anymore,
-% you may leave them out
-\terms
-term1, term2
+We consider conjunctions of equalities of the form
+$\V x_1, \ldots, \V x_n.\ \bigwedge_k x_i = x_j$, that is, conjunctions of
+equalities between variables. Some variables are \emph{flexible}, meaning they may be
+substituted for other variables. Their binder $\V x_i$ is of the form $\F \f
+x_i$ (and their are typeset in bold font). Other variables are \emph{rigid}, and
+may not be substituted. Their binder $\V x_i$ is of the form $\R x_i$. The
+problem consists in computing a substitution of \emph{flexible} variables that
+evaluate the conjunction to \li+true+. If no such substitution exists, the
+solver outputs nothing.
 
-\keywords
-keyword1, keyword2
+For instance, one may want to solve: $\R x, \R z, \F \f y.\ x = \f y \wedge z = z$.
+A solution exists: the solver outputs $\sigma = \{ \f y \mapsto x \}$ as a valid
+substitution that solves the input problem. However, if one attempts to solve:
+$\R x, \R z, \F \f y.\ x = \f y \wedge \f y = z$, the solver fails to find a
+proper substitution, and return nothing. Indeed, the first clause demands that
+$\f y$ substitutes to $x$, meaning that the second clause becomes $x = z$, which
+always evaluates to false ($x$ and $z$ are two distinct rigid variables).
 
-\section{Conjunctions of equalities}
+In proof search, rigid variables stem from the right-elimination of universal
+quantifiers, or left-elimination of existential quantifiers, which both result
+in abstract variables. When left-eliminating a universal quantifier
+(respectively right-eliminating an existential quantifier), one must provide an
+argument to the type application (resp. an existential witness). If the argument
+(resp. the witness) cannot be found on the spot, one typically uses
+\emph{flexible variables}, which allow the proof search procedure to
+\emph{defer} the choice until a later point in search (where the choice may be
+guessed). Flexible variables are thus an \emph{implementation technique}.
 
-Just the regular option monad. Lazyness is already built-in (the second
-computation is suspended).
+In order to simplify the discussion, we skip elimination rules and quantifiers
+altogether, and just assume that our input problem contains a combination of
+flexible and rigid variables. Furthermore, we assume that any instantiation of a
+flexible variable is legal. The only crucial point is that the solver outputs
+something.
 
-In this first version, we treat the simple problem of conjunction of
-equalities\dots written in an elegant style with a persistent union-find,
-functional, the option monad, because we may fail.
+(The library that we describe here has been used in the implementation of the
+Mezzo type-checker. There, all introduction and elimination rules are present;
+furthermore, the implementation uses \emph{levels} to guarantee that only sound
+instantiation choices are performed. This is, however, orthogonal to the present
+discussion; the minimal problem we consider already conveys all the important
+points, so we will skip quantifiers in the remainder of the discussion.)
 
-Lazyness already built-in, evaluation stops right after the first failing
-premise has been hit.
+Solving conjunctions of equalities requires three rules, which make up our
+logic (\fref{proof-system}). \Rule{Refl} embodies the reflexivity axiom; \Rule{And}
+highlights that the proof produces an \emph{output} $V'$ which is \emph{chained}
+from one premise to another; \Rule{Inst} embodies the ``instantiation''
+mechanism: in essence, if $y$ has not been instantiated so far, one
+may instantiate it onto any other variable and perform the substitution in order
+to prove the goal ($[\f y/x] P$ stands for ``replace $y$ with $x$ in $P$'').
 
-Goal: build a derivation tree at the same time. Requires: recording the
-premises, in order to build an application of the rule.
+\begin{figure}
+  \centering
+  \begin{mathpar}
+    \inferrule[Refl]{
+      \quad
+    }{
+      V \vdash x = x \dashv V
+    }
+
+    \inferrule[Inst]{
+      x \in V_i \\
+      V_1, \F(\f y = x), V_2 \vdash [\f y/x] P \dashv V'
+    }{
+      V_1, \F(\f y), V_2 \vdash P \dashv V'
+    }
+
+    \inferrule[And]{
+      V \vdash P \dashv V' \\
+      V' \vdash Q \dashv V''
+    }{
+      V \vdash P \wedge Q \dashv V''
+    }
+  \end{mathpar}
+  \caption{Proof rules for the first system}
+  \label{fig:proof-system}
+\end{figure}
+
+Once the proof tree has been laid out, one obtains an output (e.g. $V''$ in
+\Rule{And}) where the instantiation choices (e.g. $\F (\f x_i = x_j)$) determine
+the desired substitution (e.g. $\sigma = \{ \f x_i \mapsto x_j \}^*$).
+
+\begin{figure}
+  \centering
+  \begin{ocaml}
+type formula =                and descr =
+| Equals of var * var           | Flexible
+| And of formula * formula      | Rigid
+
+and var = P.point             and env = descr P.state
+  \end{ocaml}
+  \caption{Formulas and environments}
+  \label{fig:formulas}
+\end{figure}
+
+We implement proof search in OCaml (\fref{formulas}). The data type of formulas
+is self-explanatory. Variables are implemented as equivalence classes in a
+\emph{persistent} union-find data structure, which the module \li+P+ implements.
+The $V$ parameter in our rules is embodied by the \li+env+ type
+(``environment''), which is an input and an output to the solver. An
+\li+env+ represents a given state of the union-find data structure, that is, a
+set of equations between variables. This is another way of representing a
+substitution, i.e. this is what we want our solver to compute.
+
+The choice of a persistent union-find is irrelevant. All that matters is that we
+pick a data structure that models substitutions, and that is \emph{persistent}.
+Had we picked an explicit substitution instead of a union-find, the rest of the
+discussion would have been the same.
+
+\fref{solver} implements a solver for our minimal problem; written within the
+\li+MOption+ monad, it returns either \li+Some env+ (in case a successful
+substitution has been found), or \li+None+ if no solution exists. The solver is
+complete.
+
+\begin{figure}
+  \centering
+  \begin{ocaml}
+module MOption = struct
+  (* defines [return], [nothing] and [>>=] *)
+end
+
+let unify env v1 v2 =                         let rec solve env formula =
+  match P.find v1 env, P.find v2 env with       match formula with
+  | Flexible, Flexible                          | Equals (v1, v2) ->
+  | Flexible, Rigid ->                              unify env v1 v2
+      return (P.union v1 v2 env)                | And (f1, f2) ->
+  | Rigid, Flexible ->                              solve env f1 >>= fun env ->
+      return (P.union v2 v1 env)                    solve env f2
+  | Rigid, Rigid ->
+      if P.same v1 v2 env then
+        return env
+      else
+        nothing
+  \end{ocaml}
+  \caption{Solver for the simplified problem}
+  \label{fig:solver}
+\end{figure}
+
+The solver uses \li+MOption.>>=+ to sequence premises in the \li+And+ case. It
+doesn't keep track of premises; it just ensures (thanks to \li+>>=+) that if the
+first premise evaluates to \li+nothing+, the second premise is not evaluated,
+since it is suspended behind a \li+fun+ expression (OCaml is a strict language).
+
+\section{Building derivations}
+
+There are two shortcomings with this solver. First, the \li+unify+ sub-routine
+conflates several rules of the logic together. Indeed, the \li+return (P.union ...)+
+expression hides a combination of \Rule{Inst} and \Rule{Refl}. Second, we
+have no way to replay the proof to verify it independently. One may argue that
+in this simplified example, one can just apply the substitution to the original
+formula and verify that all the clauses are of the form $x = x$, without the
+need for a proof tree. In the general case, however, the proof tree contains the
+elimination witnesses for quantifiers; this allows one to independently verify a
+proof without relying on inference techniques, such as flexible variables.
+
+\subsection{Defining proof trees}
+
+One way to make the solver better is to make sure each step it performs
+corresponds in an obvious manner to the application of an admissible rule. To
+that effect, we define the data type of all three rules in our system, which we
+apply to the functor of \emph{proof trees} (\fref{proof-trees}).
+
+\begin{figure}
+  \centering
+\begin{ocaml}
+module type LOGIC = sig     module MakeProofTree (L: LOGIC) = struct
+  type formula                type derivation = L.formula * rule
+  type rule_name              and rule = L.rule_name * premises
+end                           and premises = Premises of derivation list
+                            end
+
+module MyLogic = struct
+  type formula = ... (* as before *)
+  type rule_name = R_And | R_Refl | R_Inst
+end
+module MyProofTree = MakeProofTree(MyLogic)
+\end{ocaml}
+  \caption{The functor of proof trees}
+  \label{fig:proof-trees}
+\end{figure}
+
+A \li+derivation+ tree is a pair of a \li+formula+ (the goal we wish the prove)
+and a \li+rule+ (that we apply in order to prove the goal). A \li+rule+ has a name
+and \li+premises+; the \li+premises+ type is simply a \li+derivation list+ (the
+\li+Premises+ constructor is here to prevent a non-constructive type
+abbreviation). When using the library, the client is expected to make sure
+that each \li+rule_name+ is paired with the proper
+number of premises (0 for \Rule{Refl}, 1 for
+\Rule{Inst} and 2 for \Rule{And}); this is not enforced by the type system.
+
+In the (simplified) sketch from \fref{proof-trees}, rule names are just constant
+constructors, since the rule parameters (such as $x$ and $\f y$ in \Rule{Inst})
+can be recovered from the \li+formula+. In the general case, the various
+constructors of \li+rule_name+ do have parameters that record how one specific
+rule was instantiated.
+
+\subsection{Proof tree combinators}
+
+We previously used the \li+>>=+ operator from the \li+MOption+ monad in order to
+chain premises (\fref{solver}). We now need a new operator, that not only
+\emph{binds} the result (i.e. stops evaluating premises after a failure, as
+before), but also \emph{records} the premises in sequence, in order to build a
+proper derivation. The former is still faithfully implemented by the option
+monad; the latter is implemented by the writer monad.
+
+Computations in the writer monad return a result (of type \li+'a+) along with a log of
+elements (of type \li+L.a+). The (usual) \li+>>=+ and \li+return+ combinators operate on
+the result part of the computation, while the (new) \li+tell+ combinator
+operates on the logging part of the computation. This \li+tell+ combinator
+appends a new element to the log. Appending elements to the log is done by way of the \li+MONOID+
+module type, which essentially demands a value for the empty log, and a function to
+append new entries into the log.
+
+In order to get a new \li+>>=+ operator that combines the features of
+the option and writer monads, we apply the \li+WriterT+ monad transformer to the
+\li+MOption+ monad (\fref{writer}) and obtain \li+MWriter+, a monad whose
+computations represent a sequence of derivations (the premises we have proved so far)
+along with a result (the \li+env+ that we chain through the premises). These
+computations are wrapped in \li+MOption.m+, that is, are wrapped within an
+\li+option+ to account for a possible proof failure.
+
+\begin{figure}
+  \centering
+\begin{ocaml}
+module WriterT (M: MONAD) (L: MONOID): sig      module L = struct
+  type 'a m = (L.a * 'a) M.m                      type a = MyProofTree.derivation list
+  val return: 'a -> 'a m                          let empty = []
+  val ( >>= ): 'a m -> ('a -> 'b m) -> 'b m       let append = List.append
+                                                end
+  val tell: L.a -> unit m
+end = ...
+
+module MWriter = WriterT(MOption)(L)
+\end{ocaml}
+  \caption{The writer monad transformer}
+  \label{fig:writer}
+\end{figure}
 

### 2015-10-12 paper/main.tex
More text in da paper.

@@
 \usepackage{minted}
-  \newminted[ocaml]{ocaml}{mathescape}
+  \newminted[ocaml]{ocaml}{mathescape,fontsize=\small}
 \usepackage{mathpartir}
@@
 
-and var = P.point             and env = descr P.state
+and var = P.point             and state = descr P.state
   \end{ocaml}
-  \caption{Formulas and environments}
+  \caption{Formulas and state}
   \label{fig:formulas}
@@
 \emph{persistent} union-find data structure, which the module \li+P+ implements.
-The $V$ parameter in our rules is embodied by the \li+env+ type
-(``environment''), which is an input and an output to the solver. An
-\li+env+ represents a given state of the union-find data structure, that is, a
-set of equations between variables. This is another way of representing a
-substitution, i.e. this is what we want our solver to compute.
+The $V$ parameter in our rules is embodied by the \li+state+ type; just
+like the $V$ parameter is chained from one premise to another (\Rule{And}),
+\li+state+ is an input and an output to the solver. Just like the $V$ parameter
+in the rules, a \li+state+ of the persistent union-find represents
+a set of equations between variables. In a sense, \li+state+ is a specific
+implementation of the theoretical $V$ parameter. It represents a
+substitution; in other words, this is what we want our solver to compute.
 
@@
 \fref{solver} implements a solver for our minimal problem; written within the
-\li+MOption+ monad, it returns either \li+Some env+ (in case a successful
+\li+MOption+ monad, it returns either \li+Some state+ (in case a successful
 substitution has been found), or \li+None+ if no solution exists. The solver is
@@
 module MOption = struct
-  (* defines [return], [nothing] and [>>=] *)
+  (* ... defines [return], [nothing] and [>>=] *)
 end
 
-let unify env v1 v2 =                         let rec solve env formula =
-  match P.find v1 env, P.find v2 env with       match formula with
+let unify state v1 v2 =                         let rec solve state formula =
+  match P.find v1 state, P.find v2 state with     match formula with
   | Flexible, Flexible                            | Equals (v1, v2) ->
-  | Flexible, Rigid ->                              unify env v1 v2
-      return (P.union v1 v2 env)                | And (f1, f2) ->
-  | Rigid, Flexible ->                              solve env f1 >>= fun env ->
-      return (P.union v2 v1 env)                    solve env f2
+  | Flexible, Rigid ->                                unify state v1 v2
+      return (P.union v1 v2 state)                | And (f1, f2) ->
+  | Rigid, Flexible ->                                solve state f1 >>= fun state ->
+      return (P.union v2 v1 state)                    solve state f2
   | Rigid, Rigid ->
-      if P.same v1 v2 env then
-        return env
+      if P.same v1 v2 state then
+        return state
       else
@@
 computations represent a sequence of derivations (the premises we have proved so far)
-along with a result (the \li+env+ that we chain through the premises). These
+along with a result (the \li+state+ that we chain through the premises). These
 computations are wrapped in \li+MOption.m+, that is, are wrapped within an
@@
 
-module MWriter = WriterT(MOption)(L)
+module M = MOption
+module MWriter = WriterT(M)(L)
 \end{ocaml}
@@
 
+A computation within this new monad has type (simplified after functor applications)
+\li+(derivation list * state) option+. It represents a given point in the proof;
+the solver is focused on a given rule, has reached a certain state, after
+proving a certain list of premises.
+
+Once all the premises have been proved, one needs to draw a horizontal line and
+reach the conclusion of the proof. That is, take the final state and the list of
+premises, and generate a \li+derivation+ that stands for the application of the
+entire rule.
+
+Contrary to the first implementation (\fref{solver}), where the working state
+and the return value of \li+solve+ both had type \li+state option+, we now
+distinguish between an \li+outcome+ (the result of a call to \li+solve+) and a
+working state (a computation in the monad).
+
+An outcome is, as we mentioned earlier, the pair of a final \li+state+ along
+with a \li+derivation+ that justifies that we reached this state. The pair is
+optional since, after all, not all formulas are provable.
+
+The type \li+outcome+ (\fref{combinators}) is parametric: it works for any
+state that the client code uses. In other words, our library is generic with
+regards to the particular \li+state+ type the client uses.
+
+\begin{figure}
+  \centering
+\begin{ocaml}
+(* This snippet is in the [MWriter(M)(L)] monad. Upon a first reading, think
+   [module M = MOption]. *)
+type 'a outcome = ('a * derivation) M.m
+
+let premise (outcome: 'a outcome): 'a m =
+  M.bind outcome (fun (state, derivation) ->
+    tell [ derivation ] >>= fun () ->
+    return state
+  )
+
+let prove (goal: goal) (rule: rule_name) (x: 'a m): 'a outcome =
+  M.(x >>= fun (premises, state) ->
+    return (state, (goal, (rule, Premises premises))))
+
+let axiom (state: 'a) (goal: goal) (axiom: rule_name): 'a outcome =
+  prove goal axiom (return state)
+
+let fail: 'a outcome =
+  M.nothing
+\end{ocaml}
+  \caption{The high-level combinators for building proof derivations}
+  \label{fig:combinators}
+\end{figure}
+
+We now have a duality between the \li+outcome+ type (the result of solving a
+sub-goal) and the \li+m+ type (a computation within the monad, i.e. a working
+state between two premises). Therefore, we introduce two high-level combinators:
+\li+premise+ and \li+prove+. The former goes from \li+outcome+ to \li+m+: it
+injects a new sub-goal as a premise of the rule we are currently trying to
+prove. The latter goes from \li+m+ to \li+outcome+: if all premises have been
+satisfied, it draws the horizontal line that builds a new node in the derivation
+tree.
+
+\begin{itemize}
+
+  \item \li+premise+ is the composition of \li+tell+, which records the
+    derivation for this sub-goal, and \li+return+, which passes the state on to
+    the next sub-goal.
+
+  \item \li+prove+ is a computation in the \li+M+ monad (here, \li+MOption+). If
+    all the premises have been satisfied, it bundles them as a new node of the
+    derivation tree. If a premise failed, then \li+x+ is \li+M.nothing+, and
+    \li+prove+ also returns a failed outcome.
+
+  \item \li+axiom+ is short-hand for a rule that requires no premises.
+
+  \item \li+fail+ is for situations where no rule applies: this is a failed outcome.
+\end{itemize}
+
+\subsection{A solver in the new style}
+
+\fref{solver2} demonstrates an implementation of \li+solve+ in the new style.
+Compared to the previous implementation (\fref{solver}):
+\begin{itemize}
+  \item \li+prove_equality+ makes it explicit which rules are applied, and
+    singles out two distinct rule applications in the flexible-rigid case;
+  \item the premises of each rule are clearly identified;
+  \item axioms and failure conditions are explicit,
+  \item the \li+And+ case is easy to review manually, to make sure that no
+    premise was forgotten.
+\end{itemize}
+This is, as mentioned previously, a minimal example that showcases the usage of
+the library. In the implementation of Mezzo, switching the core of the
+type-checker to this style revealed several bugs where premises were not
+properly chained or simply forgotten.
+
+\begin{figure}
+  \centering
+\begin{ocaml}
+let rec prove_equality (state: state) (goal: formula) (v1: var) (v2: var) =
+  let open MOption in
+  match P.find v1 state, P.find v2 state with
+  | Flexible, Flexible
+  | Flexible, Rigid ->
+      let state = P.union v1 v2 state in
+      prove goal R_Instantiate begin
+        premise (prove_equality state goal v1 v2) >>=
+        return
+      end
+  (* ... *)
+  | Rigid, Rigid ->
+      if P.same v1 v2 state then
+        axiom state goal R_Refl
+      else
+        fail
+
+let rec solve (state: state) (goal: formula): state outcome =
+  match goal with
+  | Equals (v1, v2) ->
+      prove_equality state goal v1 v2
+  | And (g1, g2) ->
+      prove goal R_And begin
+        premise (solve state g1) >>= fun state ->
+        premise (solve state g2) >>=
+        return
+      end
+\end{ocaml}
+  \caption{A solver written using the high-level combinators}
+  \label{fig:solver2}
+\end{figure}
+
+\section{Backtracking}
+
+\subsection{Limitations of the option monad}
+
+We now extend our formulas with disjunctions (\fref{choice}). A consequence
+is that we now need our base monad \li+M+ to offer a new operation; namely, one
+that, among several possible choices, picks the first one that is not a failure.
+We thus augment \li+MOption+ with a search combinator (\fref{choice}), which in
+turn allows one to implement a high-level \li+choice+ combinator for our
+library. The \li+choice+ combinator attempts to prove a \li+goal+ by trying a
+function \li+f+ on several arguments \li+a+, each of which
+is associated to a given rule. We can add one more branch to the \li+solve+
+function, which attempts to prove a disjunction by first trying a
+left-elimination (\Rule{Or-L}, \fref{proof-system2}), then a right-elimination
+(\Rule{Or-R}).
+
+\begin{figure}
+  \centering
+  \begin{mathpar}
+    \inferrule[Or-L]{
+      V \vdash P \dashv V'
+    }{
+      V \vdash P \vee Q \dashv V'
+    }
+
+    \inferrule[Or-R]{
+      V \vdash Q \dashv V'
+    }{
+      V \vdash P \vee Q \dashv V'
+    }
+  \end{mathpar}
+  \caption{New proof rules for disjunction}
+  \label{fig:proof-system2}
+\end{figure}
+
+\begin{figure}
+  \centering
+\begin{ocaml}
+(* We extend formulas with disjunctions. *)
+type formula =
+  (* ... *)
+  | Or of formula * formula
+
+(* The logic is also extend with two rules. *)
+type rule_name =
+  (* ... *)
+  | R_OrL
+  | R_OrR
+
+module MOption = struct
+  (* ... *)
+  let rec search f = function
+    | [] -> None
+    | x :: xs ->
+        match f x with
+        | Some x -> Some x
+        | None -> search f xs
+end
+
+(* Equipped with [search], we define the high-level [choice] combinator... *)
+let choice (goal: goal) (args: (rule_name * 'a) list) (f: 'a -> 'b m): 'b outcome =
+  M.search (fun (r, x) -> prove goal r (f x)) args
+
+(* ...which one uses as follows: *)
+let rec solve (state: state) (goal: formula): state outcome =
+  match goal with
+  (* ... *)
+  | Or (g1, g2) ->
+      choice goal [ R_OrL, g1; R_OrR, g2 ] (fun g ->
+        premise (solve state g) >>=
+        return
+      )
+\end{ocaml}
+  \caption{The choice combinator}
+  \label{fig:choice}
+\end{figure}
+
+The solver can now solve problems of the form $\R x, \R z, \F \f y.\ x = z \vee
+\f y = z$. It fails, however, to solve problems of the form $\R x, \R z, \F \f
+y.\ (\f y = x \vee \f y = z) \wedge \f y = z$. The reason is, the option monad
+is not powerful enough: upon finding a suitable choice in the disjunction
+case, it commits to it and drops the other one. In other words, when hitting the
+disjunction, \li+MOption+ commits to $\sigma = \{ \f y \mapsto x \}$, instead of
+keeping $\sigma = \{ \f y \mapsto z \}$ as a backup solution. Phrased yet again
+differently, we need to replace \li+MOption+ with the non-determinism monad that
+will implement \emph{backtracking}.
+
+\subsection{The exploration monad}
+
+Conceptually, we want to change our way of thinking; instead of thinking of
+\li+solve+ as a function that returns \emph{a solution}, we now think of it as a
+function that returns \emph{several possible solutions}. The state is now a
+set of states, each of which represent a possible choice.
+
+The monad of non-determinism is implemented using lists; OCaml is a strict
+language, so we write the non-determinism monad (also known as the exploration
+or backtracking monad) using lazy lists (\fref{mexplore}).
+
+\begin{figure}
+  \centering
+\begin{ocaml}
+module LL = LazyList
+module MExplore
+  type 'a m = 'a LL.t
+  let return = LL.one
+  let ( >>= ) = LL.flattenl (LL.map f x)
+  let nothing = LL.nil
+  let search f l = LL.bind (LL.of_list l) f
+end
+\end{ocaml}
+  \caption{The exploration monad}
+  \label{fig:mexplore}
+\end{figure}
+
+The reader can now go back and replace \li+module M = MOption+ with
+\li+module M = MExplore+ in \fref{writer}. The rest of the library remains
+unchanged; the \li+solve+ function (the client code) is also unchanged; and the
+combinators of the library now implement backtracking.
+
+In particular, the earlier example of $\R x, \R z, \F \f
+y.\ (\f y = x \vee \f y = z) \wedge \f y = z$ is now successfully solved by the
+library. Thanks to laziness, no extra computations occur; further solutions down
+the lazy list are only evaluated if the first ones failed.
+
+\section{Conclusion}
+
 \bibliographystyle{abbrvnat}

### 2015-10-12 paper/main.tex
Draft conclusion + formulating search with substitutions.

@@
 \newcommand{\R}{\ensuremath{\mathcal{R}}} % Rigid
-\newcommand{\f}[1]{\ensuremath{\mathbf{#1}}} % Bold
+\newcommand{\f}[1]{\ensuremath{#1^?}} % Bold
 \newcommand{\F}{\ensuremath{\mathcal{F}}} % Flexible
@@
   rules. The user calls functions such as \li+premise+, \li+prove+ or
-  \li+choice+; the library then takes care of generating a derivation tree,
-  which serves either as a witness (success) or as an error message (failure).
+  \li+choice+; the library then takes care of generating a derivation tree.
   Proof search engines written in this style enjoy: first, a one-to-one
@@
   makes manual inspection easier; second, proof witnesses ``for free'', which
-  makes a verified-validation approach easier too.
+  makes a verified, independent validation approach easier too.
 \end{abstract}
@@
 We consider conjunctions of equalities of the form
-$\V x_1, \ldots, \V x_n.\ \bigwedge_k x_i = x_j$, that is, conjunctions of
+$\bigwedge_k x_i = x_j$, that is, conjunctions of
 equalities between variables. Some variables are \emph{flexible}, meaning they may be
-substituted for other variables. Their binder $\V x_i$ is of the form $\F \f
-x_i$ (and their are typeset in bold font). Other variables are \emph{rigid}, and
-may not be substituted. Their binder $\V x_i$ is of the form $\R x_i$. The
-problem consists in computing a substitution of \emph{flexible} variables that
-evaluate the conjunction to \li+true+. If no such substitution exists, the
+substituted with other variables. They are typeset as follows: $\f y$.
+Rigid variables may not be substituted. They are typeset as follows: $x$. The
+problem consists in computing a substitution of \emph{flexible} variables so that
+the conjunction evaluates to \li+true+. If no such substitution exists, the
 solver outputs nothing.
 
-For instance, one may want to solve: $\R x, \R z, \F \f y.\ x = \f y \wedge z = z$.
+For instance, one may want to solve: $x = \f y \wedge z = z$.
 A solution exists: the solver outputs $\sigma = \{ \f y \mapsto x \}$ as a valid
 substitution that solves the input problem. However, if one attempts to solve:
-$\R x, \R z, \F \f y.\ x = \f y \wedge \f y = z$, the solver fails to find a
+$x = \f y \wedge \f y = z$, the solver fails to find a
 proper substitution, and return nothing. Indeed, the first clause demands that
@@
 flexible and rigid variables. Furthermore, we assume that any instantiation of a
-flexible variable is legal. The only crucial point is that the solver outputs
-something.
+flexible variable is legal. The only point that matters is that the solver
+computes a result which sequenced throughout the proof search.
 
 (The library that we describe here has been used in the implementation of the
-Mezzo type-checker. There, all introduction and elimination rules are present;
-furthermore, the implementation uses \emph{levels} to guarantee that only sound
-instantiation choices are performed. This is, however, orthogonal to the present
-discussion; the minimal problem we consider already conveys all the important
-points, so we will skip quantifiers in the remainder of the discussion.)
+Mezzo type-checker. There, the logic (the type system) does have all four
+introduction and elimination rules; in the implementation, we used \emph{levels}
+to guarantee that only sound instantiation choices are performed. This is,
+however, orthogonal to the present discussion; the minimal problem we consider
+already conveys all the important points, so we skip quantifiers in the
+remainder of the discussion.)
 
-Solving conjunctions of equalities requires three rules, which make up our
-logic (\fref{proof-system}). \Rule{Refl} embodies the reflexivity axiom; \Rule{And}
-highlights that the proof produces an \emph{output} $V'$ which is \emph{chained}
+\fref{proof-system} describes a proof search procedure for conjunctions of
+equalities. Rules poke at a set of variables $V$ to tell apart rigid variables from
+flexible variables. Each rule takes a substitution $\sigma$ as an input, and outputs
+an updated substitution $\sigma'$. A variable in $V$ is either a rigid ($x$),
+or a flexible ($\f y$).
+
+\Rule{Refl} embodies the reflexivity axiom; \Rule{And}
+highlights that the first sub-goal (premise) produces an \emph{output} $V'$
+which is then \emph{chained}
 from one premise to another; \Rule{Inst} embodies the ``instantiation''
-mechanism: in essence, if $y$ has not been instantiated so far, one
-may instantiate it onto any other variable and perform the substitution in order
-to prove the goal ($[\f y/x] P$ stands for ``replace $y$ with $x$ in $P$'').
+mechanism: in essence, if $\f y$ has not been instantiated so far, one
+may instantiate it onto any other variable (rigid or flexible), i.e. add a new
+entry for $\f y$ in the substitution. Finally, \Rule{Subst} may apply the
+substitution at any time in order to prove the goal.
 
@@
     }{
-      V \vdash x = x \dashv V
+      V, \sigma \vdash x = x \dashv \sigma
+    }
+
+    \inferrule[Subst]{
+      V, \sigma \vdash \sigma P \dashv \sigma'
+    }{
+      V, \sigma \vdash P \dashv \sigma'
     }
@@
     \inferrule[Inst]{
-      x \in V_i \\
-      V_1, \F(\f y = x), V_2 \vdash [\f y/x] P \dashv V'
+      x \in V \\ \f y \in V \\ \f y \not\in \sigma \\\\
+      V, \{ \f y \mapsto x \} \circ \sigma \vdash  P \dashv \sigma'
     }{
-      V_1, \F(\f y), V_2 \vdash P \dashv V'
+      V \vdash P \dashv \sigma'
     }
@@
     \inferrule[And]{
-      V \vdash P \dashv V' \\
-      V' \vdash Q \dashv V''
+      V, \sigma \vdash P \dashv \sigma' \\\\
+      V, \sigma' \vdash Q \dashv \sigma''
     }{
-      V \vdash P \wedge Q \dashv V''
+      V \vdash P \wedge Q \dashv \sigma''
     }
   \end{mathpar}
-  \caption{Proof rules for the first system}
+  \caption{Semi-algorithmic proof rules}
   \label{fig:proof-system}
@@
 
-Once the proof tree has been laid out, one obtains an output (e.g. $V''$ in
-\Rule{And}) where the instantiation choices (e.g. $\F (\f x_i = x_j)$) determine
-the desired substitution (e.g. $\sigma = \{ \f x_i \mapsto x_j \}^*$).
+Once the proof tree has been laid out, one obtains an output (e.g. $\sigma''$ in
+\Rule{And}). Since flexible variables may instantiate onto other flexible
+variables, one may need to apply the substitution several times. The reflexive-transitive
+closure of $\sigma$, i.e. $\sigma^*$, is thus the desired result of the proof
+search.
 
@@
 
-The solver can now solve problems of the form $\R x, \R z, \F \f y.\ x = z \vee
-\f y = z$. It fails, however, to solve problems of the form $\R x, \R z, \F \f
-y.\ (\f y = x \vee \f y = z) \wedge \f y = z$. The reason is, the option monad
+The solver can now solve problems of the form $x = z \vee
+\f y = z$. It fails, however, to solve problems of the form $(\f y = x \vee \f y = z) \wedge \f y = z$. The reason is, the option monad
 is not powerful enough: upon finding a suitable choice in the disjunction
@@
 function that returns \emph{several possible solutions}. The state is now a
-set of states, each of which represent a possible choice.
+set of states, each of which represent a path in the search tree of derivation
+trees.
 
@@
 
-In particular, the earlier example of $\R x, \R z, \F \f
-y.\ (\f y = x \vee \f y = z) \wedge \f y = z$ is now successfully solved by the
+In particular, the earlier example of $(\f y = x \vee \f y = z) \wedge \f y = z$ is now successfully solved by the
 library. Thanks to laziness, no extra computations occur; further solutions down
@@
 
+We presented a support library for writing a proof search engine using backtracking in
+any given logic; indeed, the library is parameterized by: the type of formulas; the type of
+rule applications; the internal state type of the client. By merely using the
+combinators of the library, the client gets derivations built for free; this allows
+a separate verifier to independently check the steps required to prove the
+formula. By opting into the library, the client gets to rewrite their code in a
+new syntactic style that makes rule application explicits, forbids ``bundled''
+applications of multiple rules at the same time and clearly lays out the
+premises required to prove a judgement. Since the code resembles the logical
+rules, mistakes are easier to spot.
+
+The logic presented in this paper is as simple as it gets. It does, however,
+highlight the main concepts. A version of this library is used in the core of
+Mezzo's type-checker. The version of the library used in Mezzo also builds
+failed derivations; these failed derivations stop at the first failed premise
+or, in case of a choice, list all the failed attempts. We have not yet explained
+this last feature as a clean combination of monads and operators, but hope to do
+so in the near future.
+
 \bibliographystyle{abbrvnat}

### 2015-10-12 paper/main.tex
Better motivation + qed combinator.

@@
   % command clickable in a PDF.
-  \let\TirName\textsc
-  \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
-  \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
-  \let\DefRule\RefTirName
-  \let\Rule\DefTirName
+  % \let\TirName\textsc
+  % \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
+  % \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
+  % \let\DefRule\RefTirName
+  % \let\Rule\DefTirName
+  \let\Rule\textsc
 
 \newcommand{\fref}[1]{Figure~\ref{fig:#1}}
+\newcommand{\sref}[1]{Section~\ref{sec:#1}}
 \newcommand{\R}{\ensuremath{\mathcal{R}}} % Rigid
@@
 
-\section{A minimal problem}
+\section{Theory and practice}
 
-We consider conjunctions of equalities of the form
-$\bigwedge_k x_i = x_j$, that is, conjunctions of
-equalities between variables. Some variables are \emph{flexible}, meaning they may be
-substituted with other variables. They are typeset as follows: $\f y$.
-Rigid variables may not be substituted. They are typeset as follows: $x$. The
-problem consists in computing a substitution of \emph{flexible} variables so that
-the conjunction evaluates to \li+true+. If no such substitution exists, the
-solver outputs nothing.
+\subsection{A minimal theory}
 
-For instance, one may want to solve: $x = \f y \wedge z = z$.
-A solution exists: the solver outputs $\sigma = \{ \f y \mapsto x \}$ as a valid
-substitution that solves the input problem. However, if one attempts to solve:
-$x = \f y \wedge \f y = z$, the solver fails to find a
-proper substitution, and return nothing. Indeed, the first clause demands that
-$\f y$ substitutes to $x$, meaning that the second clause becomes $x = z$, which
-always evaluates to false ($x$ and $z$ are two distinct rigid variables).
+We are concerned with proving the validity of logical formulas; that is, with
+writing a search procedure that determines whether a given goal is satisfiable.
+To get started, we consider a system made up of conjunctions of equalities,
+along with existential quantifiers. Any free variables are assumed to be
+universally quantified. For instance, one may want to prove the following
+formula:
 
-In proof search, rigid variables stem from the right-elimination of universal
-quantifiers, or left-elimination of existential quantifiers, which both result
-in abstract variables. When left-eliminating a universal quantifier
-(respectively right-eliminating an existential quantifier), one must provide an
-argument to the type application (resp. an existential witness). If the argument
-(resp. the witness) cannot be found on the spot, one typically uses
-\emph{flexible variables}, which allow the proof search procedure to
-\emph{defer} the choice until a later point in search (where the choice may be
-guessed). Flexible variables are thus an \emph{implementation technique}.
+\begin{equation}
+  \exists y.\ x = y
+  \label{ex1}
+\end{equation}
 
-In order to simplify the discussion, we skip elimination rules and quantifiers
-altogether, and just assume that our input problem contains a combination of
-flexible and rigid variables. Furthermore, we assume that any instantiation of a
-flexible variable is legal. The only point that matters is that the solver
-computes a result which sequenced throughout the proof search.
+In order to show the validity of this judgement, one will build a proof
+derivation using the rules from the logic, shown in \fref{logic} ($[x/y]P$ means
+``substitute $x$ with $y$ in $P$''). For instance, proving Equation \ref{ex1}
+requires applying \Rule{ExistsE}, then \Rule{Refl}.
 
-(The library that we describe here has been used in the implementation of the
-Mezzo type-checker. There, the logic (the type system) does have all four
-introduction and elimination rules; in the implementation, we used \emph{levels}
-to guarantee that only sound instantiation choices are performed. This is,
-however, orthogonal to the present discussion; the minimal problem we consider
-already conveys all the important points, so we skip quantifiers in the
-remainder of the discussion.)
+\begin{figure}
+  \centering
+  \begin{mathpar}
+    \inferrule[Refl]{
+      \quad
+    }{
+      x = x
+    }
 
-\fref{proof-system} describes a proof search procedure for conjunctions of
-equalities. Rules poke at a set of variables $V$ to tell apart rigid variables from
-flexible variables. Each rule takes a substitution $\sigma$ as an input, and outputs
-an updated substitution $\sigma'$. A variable in $V$ is either a rigid ($x$),
-or a flexible ($\f y$).
+    \inferrule[And]{
+      P \\ Q
+    }{
+      P \wedge Q
+    }
 
-\Rule{Refl} embodies the reflexivity axiom; \Rule{And}
-highlights that the first sub-goal (premise) produces an \emph{output} $V'$
-which is then \emph{chained}
-from one premise to another; \Rule{Inst} embodies the ``instantiation''
-mechanism: in essence, if $\f y$ has not been instantiated so far, one
-may instantiate it onto any other variable (rigid or flexible), i.e. add a new
-entry for $\f y$ in the substitution. Finally, \Rule{Subst} may apply the
-substitution at any time in order to prove the goal.
+    \inferrule[ExistsE]{
+      [x/y]P
+    }{
+      \exists y.\ P
+    }
+  \end{mathpar}
+  \caption{A simple logic}
+  \label{fig:logic}
+\end{figure}
+
+These rules embody the Truth of our logic, i.e. an omniscient reader may use
+them to show with absolute certainty that a given formula is true.
+%
+However, if one wants to algorithmically determine whether a given formula is
+true, \Rule{ExistsE} is useless. Indeed, unless the algorithm (solver) is
+equipped with superpowers, it cannot magically guess, out of the blue, a
+suitable $x$ in \Rule{ExistsE} that will ensure the remainder of the derivation
+succeeds. To put it another way, $x$ is a free variable (a parameter) of
+\Rule{ExistsE}; the whole point of writing a proof search algorithm is to 1)
+find that \Rule{ExistsE} is the right rule to apply, and 2) find that $x$ is a
+suitable value for instantiating $y$, because it will make $y = x$ succeed.
+
+Hence, in order to build a \emph{search procedure} for that logic, one will
+use another set of \emph{algorithmic} rules, which hopefully enjoy:
+\begin{description}
+  \item[soundness]: if the algorithmic rules succeed, then there exists a
+    derivation in the logic that proves the validity of the original formula,
+    and
+  \item[completeness]: if the algorithmic rules fail, then there exists no
+    derivation in the logic that would prove the validity of the original
+    formula.
+\end{description}
+
+For instance, in our logic of existentially-quantified conjunctions of
+equalities, one may want to use the algorithmic rules from
+\fref{proof-system}. These rules differ from \fref{logic} in that they are
+algorithmic; they take an input and return an output.
 
@@
     }
+
+    \inferrule[ExistsE]{
+      V \uplus \f y, \sigma \vdash P \dashv \sigma'
+    }{
+      V, \sigma \vdash \exists y.\ P \dashv \sigma'_{|V}
+    }
   \end{mathpar}
-  \caption{Semi-algorithmic proof rules}
+  \caption{Algorithmic proof rules}
   \label{fig:proof-system}
@@
 
-Once the proof tree has been laid out, one obtains an output (e.g. $\sigma''$ in
-\Rule{And}). Since flexible variables may instantiate onto other flexible
-variables, one may need to apply the substitution several times. The reflexive-transitive
-closure of $\sigma$, i.e. $\sigma^*$, is thus the desired result of the proof
-search.
+In particular, in order to determine suitable values for the $x$ parameter in
+\Rule{ExistsE}, the implementation reasons in terms of substitutions. $V$ is a
+set of variables which may be substituted (recall that free variables are
+considered universally quantified, hence not eligible for substitution);
+variables that may be substituted are typeset as $\f y$.
+The algorithm has internal state, that is, it carries a substitution $\sigma$.
+Upon hitting an existential quantifier $\f y$, the algorithmic rules
+\emph{open} $\f y$ and mark it as eligible for substitution. Later on (for
+instance, upon hitting $\f y = x$), the algorithm may pick a
+substitution for $\f y$ using \Rule{Inst}. A substitution may be applied at any
+time (\Rule{Subst}). The preconditions of \Rule{Inst} guarantee that the
+algorithm makes at most one choice for instantiating $\f y$.
+
+In other words, the algorithmic rules \emph{defer} the \emph{instantiation} of
+the existential quantifier until the shape of the proof obligation gives us a
+\emph{hint} as to what exactly this instantiation should be. This
+\emph{implementation technique} is known as \emph{flexible variables}.
+
+The new algorithmic rules differ from the original logical rules significantly;
+first, there are five rules for the algorithmic system, compared to just three
+for the logical system. Second, these five rules do not map trivially to their
+counterparts in the logical system. Third, these rules are only
+algorithmic; the implementation that we are about to roll out uses an
+optimized representation for substitutions (union-find), meaning that one not
+only needs to check that the algorithmic rules are faithful to the logical
+rules, but also that the implementation is faithful to the algorithmic rules.
+
+This paper presents a library that allows one to write an implementation of the
+algorithmic rules while automatically generating a derivation. The library
+forces the client code to lay out premises, rule applications and
+instantiations. The level of detail of the resulting derivation is left up to
+the client code; they may wish to record the very compact rules from the logic,
+or record more proof steps using the algorithmic rules. In any case, the
+derivation serves as a proof witness; if the user wishes to do so, they can
+write a validator that takes the witness and verifies that the derivation is,
+indeed, correct.
+
+The library has been used, in a preliminary form, to implement the core of the
+Mezzo type-checker. This paper presents a cleaned-up, isolated version of this
+library.
+
+\subsection{An implementation with flexible variables and union-find}
+
+The logic we present is a much simplified version of the logic (type system)
+of Mezzo. In particular, this paper only mentions the right-exists quantifier;
+Mezzo has all four possible combinations of left/right exists/forall. Generally,
+in proof search, \emph{flexible variables} stem from the right-elimination of
+existential quantifiers, or the left-elimination of universal quantifiers. The
+right-elimination of universal quantifiers, or the left-elimination of
+existential quantifiers give rise to universally-quantified variables, which are
+called \emph{rigid}.
+
+In order to simplify the problem, we assume that all existential variables have
+been introduced as flexible variables already. That way, we won't be
+sidetracked, talking about binders and the respective merits of De Bruijn
+\emph{vs.} locally nameless. Furthermore, we assume that all instantiations of
+flexible variables are legal. This is not true in general: for instance, if the
+goal is $\forall x, \exists \f y, \forall z.\ P$, picking $\f y = z$ makes no
+sense. Mezzo forbids this choice using \emph{levels}; in the present document,
+we skip this discussion altogether and assume that ``all is well''. Finally,
+although in a general setting, several rules may trigger for a given goal (this
+is the case in Mezzo), the algorithmic set of rules we use is syntax-driven: the
+syntactic shape of the goal determines which rule should be applied.
+
+We thus restrict our formulas to conjunctions of equalities between variables.
+The plan is to write a solver that takes,
+as an input, a formula, and outputs a valid substitution, if any. That is,
+write an algorithm that abides by the rules from \fref{proof-system}.
+For instance, one may want to solve: $x = \f y \wedge z = z$.
+A solution exists: the solver outputs $\sigma = \{ \f y \mapsto x \}$ as a valid
+substitution that solves the input problem. However, if one attempts to solve:
+$x = \f y \wedge \f y = z$, the solver fails to find a
+proper substitution, and returns nothing. Indeed, the first clause demands that
+$\f y$ substitutes to $x$, meaning that the second clause becomes $x = z$, which
+always evaluates to false ($x$ and $z$ are two distinct rigid variables).
+
+Once the algorithm has run, we obtain an output substitution $\sigma$. One can, if they
+wish to do so, take the reflexive-transitive closure $\sigma^*$, and apply it to
+a flexible variable (say, $\f y$) to recover the parameter of \Rule{ExistsE}
+that should be used in the logical rules (here, $x$). This is all very informal
+-- the point of the subsequent sections is to formalize the claim that the proof
+search algorithm produces a proof witness.
 
@@
 
-We implement proof search in OCaml (\fref{formulas}). The data type of formulas
+We implement proof search in OCaml (\fref{formulas}); we do not use explicit
+substitutions, but rather an optimized representation based on a union-find data
+structure. The data type of formulas
 is self-explanatory. Variables are implemented as equivalence classes in a
 \emph{persistent} union-find data structure, which the module \li+P+ implements.
-The $V$ parameter in our rules is embodied by the \li+state+ type; just
-like the $V$ parameter is chained from one premise to another (\Rule{And}),
-\li+state+ is an input and an output to the solver. Just like the $V$ parameter
+The $V, \sigma$ parameters in our rules are embodied by the \li+state+ type; just
+like the $\sigma$ parameter is chained from one premise to another (\Rule{And}),
+\li+state+ is an input and an output to the solver. Just like the $\sigma$ parameter
 in the rules, a \li+state+ of the persistent union-find represents
 a set of equations between variables. In a sense, \li+state+ is a specific
-implementation of the theoretical $V$ parameter. It represents a
+implementation of the theoretical $\sigma$ parameter. It represents a
 substitution; in other words, this is what we want our solver to compute.
 
-The choice of a persistent union-find is irrelevant. All that matters is that we
+The choice of a union-find is irrelevant. All that matters is that we
 pick a data structure that models substitutions, and that is \emph{persistent}.
@@
 There are two shortcomings with this solver. First, the \li+unify+ sub-routine
-conflates several rules of the logic together. Indeed, the \li+return (P.union ...)+
+conflates several rules together. Indeed, the \li+return (P.union ...)+
 expression hides a combination of \Rule{Inst} and \Rule{Refl}. Second, we
 have no way to replay the proof to verify it independently. One may argue that
-in this simplified example, one can just apply the substitution to the original
+in this simplified example, the outputs substitution \emph{is} the proof
+witness: one can just apply the substitution to the original
 formula and verify that all the clauses are of the form $x = x$, without the
 need for a proof tree. In the general case, however, the proof tree contains the
-elimination witnesses for quantifiers; this allows one to independently verify a
-proof without relying on inference techniques, such as flexible variables.
+\Rule{ExistsE} rule, and proof witnesses are attached to arbitrary nodes of the
+tree. We thus need to build a properly annotated proof tree in the general case.
 
@@
 
+We record applications of \Rule{Inst}, \Rule{Refl} and \Rule{And}. This produces
+a derivation tree that makes sure that the algorithm follows the algorithmic
+rules from \fref{proof-system}. \sref{generate-logical} shows how to generate a
+slightly different tree that matches the rules from \fref{logic}.
+
 \begin{figure}
@@
 \begin{ocaml}
+(* These two modules belong to the library. *)
 module type LOGIC = sig     module MakeProofTree (L: LOGIC) = struct
@@
 
+(* This is the client code using modules from the library. *)
 module MyLogic = struct
@@
 \end{ocaml}
-  \caption{The functor of proof trees}
+  \caption{The functor of proof trees (library and client code)}
   \label{fig:proof-trees}
@@
 \end{ocaml}
-  \caption{The writer monad transformer}
+  \caption{The writer monad transformer (library code)}
   \label{fig:writer}
@@
 
-Once all the premises have been proved, one needs to draw a horizontal line and
+We provide a convenience \li+qed+ combinator: once one has obtained the final
+state, it pairs the state with the name of the rule we want to conclude with. It
+makes the implementation of \li+solve+ (\fref{solver2}) more elegant.
+
+Once all the premises have been proven, one needs to draw a horizontal line and
 reach the conclusion of the proof. That is, take the final state and the list of
@@
 with a \li+derivation+ that justifies that we reached this state. The pair is
-optional since, after all, not all formulas are provable.
+wrapped in \li+M.t+ (here, \li+option+): proving a formula may fail.
 
@@
 
-let prove (goal: goal) (rule: rule_name) (x: 'a m): 'a outcome =
-  M.(x >>= fun (premises, state) ->
+let prove (goal: goal) (x: ('a * rule_name) m): 'a outcome =
+  M.(x >>= fun (premises, (state, rule)) ->
     return (state, (goal, (rule, Premises premises))))
@@
 let axiom (state: 'a) (goal: goal) (axiom: rule_name): 'a outcome =
-  prove goal axiom (return state)
+  prove goal (return (state, axiom))
+
+let qed r e =
+  return (e, r)
 
@@
 \end{ocaml}
-  \caption{The high-level combinators for building proof derivations}
+  \caption{The high-level combinators for building proof derivations (library
+  code)}
   \label{fig:combinators}
@@
       let state = P.union v1 v2 state in
-      prove goal R_Instantiate begin
+      prove goal begin
         premise (prove_equality state goal v1 v2) >>=
-        return
+        qed R_Instantiate
       end
@@
   | And (g1, g2) ->
-      prove goal R_And begin
+      prove goal begin
         premise (solve state g1) >>= fun state ->
         premise (solve state g2) >>=
-        return
+        qed R_And
       end
 \end{ocaml}
-  \caption{A solver written using the high-level combinators}
+  \caption{A solver written using the high-level combinators (client code)}
   \label{fig:solver2}
@@
 
-(* The logic is also extend with two rules. *)
+(* The logic is also extended with two rules. *)
 type rule_name =
@@
 
-(* Equipped with [search], we define the high-level [choice] combinator... *)
-let choice (goal: goal) (args: (rule_name * 'a) list) (f: 'a -> 'b m): 'b outcome =
-  M.search (fun (r, x) -> prove goal r (f x)) args
+(* Equipped with [search], we define the [choice] library combinator... *)
+let choice (goal: goal) (args: 'a list) (f: 'a -> ('b * rule_name) m): 'b outcome =
+  M.search (fun x -> prove goal (f x)) args
 
@@
   | Or (g1, g2) ->
-      choice goal [ R_OrL, g1; R_OrR, g2 ] (fun g ->
+      choice goal [ R_OrL, g1; R_OrR, g2 ] (fun (r, g) ->
         premise (solve state g) >>=
-        return
+        qed r
       )
 \end{ocaml}
-  \caption{The choice combinator}
+  \caption{The choice combinator (library and client code)}
   \label{fig:choice}
@@
 
-The solver can now solve problems of the form $x = z \vee
-\f y = z$. It fails, however, to solve problems of the form $(\f y = x \vee \f y = z) \wedge \f y = z$. The reason is, the option monad
-is not powerful enough: upon finding a suitable choice in the disjunction
-case, it commits to it and drops the other one. In other words, when hitting the
-disjunction, \li+MOption+ commits to $\sigma = \{ \f y \mapsto x \}$, instead of
-keeping $\sigma = \{ \f y \mapsto z \}$ as a backup solution. Phrased yet again
-differently, we need to replace \li+MOption+ with the non-determinism monad that
-will implement \emph{backtracking}.
+The solver can now solve problems of the form $x = z \vee \f y = z$. It fails,
+however, to solve problems of the form $(\f y = x \vee \f y = z) \wedge \f y =
+z$. The reason is, the option monad is not powerful enough: upon finding a
+suitable choice in the disjunction case, it commits to it and drops the other
+one. In other words, when hitting the disjunction, \li+MOption+ commits to
+$\sigma = \{ \f y \mapsto x \}$, instead of keeping $\sigma = \{ \f y \mapsto z
+\}$ as a backup solution. Phrased yet again differently, we need to replace
+\li+MOption+ with the non-determinism monad that will implement
+\emph{backtracking}.
 
@@
 
+\section{Generating a proof tree using the logical rules}
+\label{sec:generate-logical}
+
 \section{Conclusion}

### 2015-10-13 paper/main.tex
More chit-chat in the paper.

@@
 (* These two modules belong to the library. *)
-module type LOGIC = sig     module MakeProofTree (L: LOGIC) = struct
+module type LOGIC = sig     module Derivations.Make (L: LOGIC) = struct
   type formula                type derivation = L.formula * rule
@@
 end
-module MyProofTree = MakeProofTree(MyLogic)
+module MyDerivations = Derivations.Make(MyLogic)
 \end{ocaml}
@@
 module WriterT (M: MONAD) (L: MONOID): sig      module L = struct
-  type 'a m = (L.a * 'a) M.m                      type a = MyProofTree.derivation list
+  type 'a m = (L.a * 'a) M.m                      type a = Derivations.derivation list
   val return: 'a -> 'a m                          let empty = []
@@
 
-\section{Generating a proof tree using the logical rules}
+\section{Full example: quantifiers and derivations of logical rules}
 \label{sec:generate-logical}
+
+The final section is concerned with adding quantifiers to the language, and
+generating derivation trees that reflect the rules of the \emph{logic} rather
+than the rules of the \emph{algorithm}. The former is useful for debugging the
+inner workings of the algorithm; the latter is useful for producing a proof
+certificate.
+
+\begin{figure}
+  \centering
+\begin{ocaml}
+(* Update of the [MyLogic] module. *)
+type formula =
+  (* ... *)
+  | Exists of string * formula
+  | Forall of string * formula
+
+type rule_name =
+  | R_And
+  | R_Refl of atom
+  | R_OrL
+  | R_OrR
+  | R_ExistsE of atom
+  | R_ForallE
+
+(* Update of the [Derivations] module. *)
+type derivation =
+  goal * rule
+
+and goal =
+  L.state * L.formula
+
+and (* ... *)
+
+(* Update of the [Combinators] module. *)
+let prove (goal: Logic.formula) (x: ('a * rule_name) m): 'a outcome =
+  M.(x >>= fun (premises, (env, rule)) ->
+    return (env, ((env, goal), (rule, Premises premises))))
+
+(* Update of the client. *)
+let rec solve (state: state) (goal: formula): state outcome =
+  (* ... *)
+  | Exists (atom, g) ->
+      let var, g, state = open_flexible state atom g in
+      let var = assert_open var in
+      prove goal begin
+        premise (solve state g) >>= fun state ->
+        qed (R_ExistsE (name var state)) state
+      end
+\end{ocaml}
+  \caption{Dealing with quantifiers}
+  \label{fig:quantifiers}
+\end{figure}
+
+We augment the data type of formulas with quantifiers; we replace the type of
+rules with the rules from the logic. Furthermore, we demand that the \Rule{Refl}
+and \Rule{ExistsE} rules be annotated with their argument.
+
+Bound variables are globally-unique atoms (strings); open variables are
+equivalence classes of the union-find, as before.
+
+Now, a \li+goal+ is no longer just a formula: it is a formula interpreted in a
+given state. Indeed, since we now open binders and substitute bound variables
+for open variables, reading a formula only makes sense in a state where the open
+variables of the formula are properly bound in the environment.
+
+We thus change the \li+prove+ combinator to record the state upon creating
+a new node in the derivation tree. More precisely, the \li+prove+ combinator
+records the state after the premises have been satisfied. This is exactly what
+is needed to conclude a rule; for instance, in the \li+Exists+ case, the solver
+prods the union-find state to discover the instantiation choice that
+\emph{has been made} for the existentially-quantified variable, and records it
+in the proof tree.
+
+Here is a pretty-printed version of a derivation tree.
+
+\begin{verbatim}
+prove ∀x. ∀z. ∃y. (y = x \/ y = z) /\ y = z using [forall]
+| prove ∀z. ∃y. (y = x \/ y = z) /\ y = z using [forall]
+| | prove ∃y. (y = x \/ y = z) /\ y = z using [exists[z]]
+| | | prove (z = x \/ z = z) /\ z = z using [/\]
+| | | | prove z = x \/ z = z using [\/_r]
+| | | | | prove z = z using axiom [refl[z]]
+| | | | prove z = z using axiom [refl[z]]
+\end{verbatim}
 

### 2015-10-13 paper/main.tex
Cleanups.

@@
 
+\usepackage{enumerate}
 \usepackage{tabularx}
@@
   % command clickable in a PDF.
-  % \let\TirName\textsc
-  % \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
-  % \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
-  % \let\DefRule\RefTirName
-  % \let\Rule\DefTirName
-  \let\Rule\textsc
+  \let\TirName\textsc
+  \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
+  \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
+  \let\DefRule\RefTirName
+  \let\Rule\DefTirName
+  % \let\Rule\textsc
 
@@
   We present the proof search monad, a set of combinators that allows one to
-  write a proof search engine in a style that resembles closely the inference
-  rules. The user calls functions such as \li+premise+, \li+prove+ or
+  write a proof search engine in a style that resembles the formal rules
+  closely. The user calls functions such as \li+premise+, \li+prove+ or
   \li+choice+; the library then takes care of generating a derivation tree.
   Proof search engines written in this style enjoy: first, a one-to-one
-  correspondence between the implementation and the theoretical rules, which
+  correspondence between the implementation and the derivation rules, which
   makes manual inspection easier; second, proof witnesses ``for free'', which
@@
 \section{Theory and practice}
+\label{sec:intro}
 
@@
 
-In order to show the validity of this judgement, one will build a proof
-derivation using the rules from the logic, shown in \fref{logic} ($[x/y]P$ means
-``substitute $x$ with $y$ in $P$''). For instance, proving Equation \ref{ex1}
-requires applying \Rule{ExistsE}, then \Rule{Refl}.
+In order to show the validity of this judgement, one usually builds a proof
+derivation using rules from the logic. In our case, the rules are given in
+\fref{logic}; $[x/y]P$ means ``substitute $x$ with $y$ in $P$''. For
+instance, proving Equation \ref{ex1} requires applying \Rule{ExistsE}, then
+\Rule{Refl}.
 
@@
 For instance, in our logic of existentially-quantified conjunctions of
-equalities, one may want to use the algorithmic rules from
-\fref{proof-system}. These rules differ from \fref{logic} in that they are
-algorithmic; they take an input and return an output.
+equalities, one may want to use the rules from \fref{proof-system}. These rules
+differ from \fref{logic} in that they are algorithmic; they take an input and
+return an output.
 
@@
 Upon hitting an existential quantifier $\f y$, the algorithmic rules
-\emph{open} $\f y$ and mark it as eligible for substitution. Later on (for
-instance, upon hitting $\f y = x$), the algorithm may pick a
+\emph{open} $\f y$ and mark it as eligible for substitution (\Rule{ExistsE}).
+Later on (for instance, upon hitting $\f y = x$), the algorithm may pick a
 substitution for $\f y$ using \Rule{Inst}. A substitution may be applied at any
@@
 In other words, the algorithmic rules \emph{defer} the \emph{instantiation} of
-the existential quantifier until the shape of the proof obligation gives us a
-\emph{hint} as to what exactly this instantiation should be. This
-\emph{implementation technique} is known as \emph{flexible variables}.
+the existential quantifier until some sub-goal, later on, gives us a \emph{hint}
+as to what exactly this instantiation should be. This \emph{implementation
+technique} is known as \emph{flexible variables}.
 
@@
 for the logical system. Second, these five rules do not map trivially to their
-counterparts in the logical system. Third, these rules are only
-algorithmic; the implementation that we are about to roll out uses an
-optimized representation for substitutions (union-find), meaning that one not
-only needs to check that the algorithmic rules are faithful to the logical
-rules, but also that the implementation is faithful to the algorithmic rules.
+counterparts in the logical system. Third, these rules are still very much
+abstract; the implementation that we are about to roll out uses an
+optimized representation for substitutions (union-find) that is not formalized
+in \fref{proof-system}. Phrased differently, one not only needs to check that
+the algorithmic rules are faithful to the proof rules, but also that the
+implementation itself is faithful to the algorithmic rules.
 
@@
 instantiations. The level of detail of the resulting derivation is left up to
-the client code; they may wish to record the very compact rules from the logic,
-or record more proof steps using the algorithmic rules. In any case, the
-derivation serves as a proof witness; if the user wishes to do so, they can
-write a validator that takes the witness and verifies that the derivation is,
-indeed, correct.
+the client code; the user may wish to record a proof derivation using the proof
+rules, or record a trace of the algorithm using the algorithmic rules.  In any
+case, the derivation serves as a witness; in the case of a proof derivation, a
+validator may certify that the proof is valid, while in the case of an
+algorithmic trace, the user may verify the algorithm, or inspect the trace for
+debugging or feedback purposes.
 
@@
 Mezzo type-checker. This paper presents a cleaned-up, isolated version of this
-library.
+library that exposes a proper interface using monads and domain-specific
+combinators.
 
@@
 
-The logic we present is a much simplified version of the logic (type system)
-of Mezzo. In particular, this paper only mentions the right-exists quantifier;
-Mezzo has all four possible combinations of left/right exists/forall. Generally,
-in proof search, \emph{flexible variables} stem from the right-elimination of
-existential quantifiers, or the left-elimination of universal quantifiers. The
+The logic we present is a much simplified version of the logic (type system) of
+Mezzo. In this document, we only mention the right-exists quantifier. General
+systems such as Mezzo have all four possible combinations of left/right
+exists/forall.
+%
+The right-elimination of existential quantifiers, or the left-elimination of
+universal quantifiers gives \emph{flexible variables}, while the
 right-elimination of universal quantifiers, or the left-elimination of
-existential quantifiers give rise to universally-quantified variables, which are
-called \emph{rigid}.
+existential quantifiers gives universally-quantified variables, also called
+\emph{rigid variables}.
 
@@
 a flexible variable (say, $\f y$) to recover the parameter of \Rule{ExistsE}
-that should be used in the logical rules (here, $x$). This is all very informal
--- the point of the subsequent sections is to formalize the claim that the proof
-search algorithm produces a proof witness.
+that should be used in the logical rules (here, $x$). This way of checking
+correctness is not satisfactory and does not scale if nested quantifiers appear
+in the goal; the point of the subsequent sections is to make sure the search
+algorithm produces a proper proof witness (derivation) that has a tree-like
+structure and does not leak implementation details (such as substitutions).
 
@@
 
-We implement proof search in OCaml (\fref{formulas}); we do not use explicit
-substitutions, but rather an optimized representation based on a union-find data
-structure. The data type of formulas
-is self-explanatory. Variables are implemented as equivalence classes in a
+We implement proof search in OCaml (\fref{formulas}); we implement substitutions
+using a union-find data structure. The data type of formulas
+is self-explanatory. Variables are implemented as equivalence classes in the
 \emph{persistent} union-find data structure, which the module \li+P+ implements.
@@
 in the rules, a \li+state+ of the persistent union-find represents
-a set of equations between variables. In a sense, \li+state+ is a specific
-implementation of the theoretical $\sigma$ parameter. It represents a
-substitution; in other words, this is what we want our solver to compute.
+a set of equations between variables. The algorithmic rules mentioned a
+theoretical $\sigma$ parameter; the \li+state+ is our specific implementation
+choice.
 
-The choice of a union-find is irrelevant. All that matters is that we
-pick a data structure that models substitutions, and that is \emph{persistent}.
-Had we picked an explicit substitution instead of a union-find, the rest of the
-discussion would have been the same.
+The choice of a union-find (as opposed to explicit substitutions) is irrelevant.
+All that matters is that we pick a data structure that models substitutions, and
+that the structure be \emph{persistent}.
 
@@
 
-let unify state v1 v2 =                         let rec solve state formula =
+let unify state v1 v2: state =                  let rec solve state formula: state =
   match P.find v1 state, P.find v2 state with     match formula with
@@
 \section{Building derivations}
+\label{sec:derivations}
 
@@
 We record applications of \Rule{Inst}, \Rule{Refl} and \Rule{And}. This produces
-a derivation tree that makes sure that the algorithm follows the algorithmic
-rules from \fref{proof-system}. \sref{generate-logical} shows how to generate a
-slightly different tree that matches the rules from \fref{logic}.
+a derivation tree (algorithm trace) that makes sure that the algorithm follows
+the algorithmic rules from \fref{proof-system}. \sref{generate-logical} shows
+how to generate a different, more compact tree that matches the rules from
+\fref{logic}.
 
@@
 constructors, since the rule parameters (such as $x$ and $\f y$ in \Rule{Inst})
-can be recovered from the \li+formula+. In the general case, the various
-constructors of \li+rule_name+ do have parameters that record how one specific
-rule was instantiated.
+can be recovered from the \li+formula+. In the general case
+(\sref{generate-logical}), the various constructors of \li+rule_name+ do have
+parameters that record how one specific rule was instantiated.
 
@@
 the result part of the computation, while the (new) \li+tell+ combinator
-operates on the logging part of the computation. This \li+tell+ combinator
-appends a new element to the log. Appending elements to the log is done by way of the \li+MONOID+
-module type, which essentially demands a value for the empty log, and a function to
-append new entries into the log.
+operates on the logging part of the computation. The \li+tell+ combinator
+appends a new element to the log; this is done by way of the \li+MONOID+ module
+type, which essentially demands a value for the \li+empty+ log, and a function
+to \li+append+ new entries into the log.
 
@@
 the option and writer monads, we apply the \li+WriterT+ monad transformer to the
-\li+MOption+ monad (\fref{writer}) and obtain \li+MWriter+, a monad whose
-computations represent a sequence of derivations (the premises we have proved so far)
-along with a result (the \li+state+ that we chain through the premises). These
-computations are wrapped in \li+MOption.m+, that is, are wrapped within an
-\li+option+ to account for a possible proof failure.
+\li+MOption+ monad (\fref{writer}) and obtain \li+MWriter+.
 
@@
 
-A computation within this new monad has type (simplified after functor applications)
-\li+(derivation list * state) option+. It represents a given point in the proof;
-the solver is focused on a given rule, has reached a certain state, after
-proving a certain list of premises.
-
-We provide a convenience \li+qed+ combinator: once one has obtained the final
-state, it pairs the state with the name of the rule we want to conclude with. It
-makes the implementation of \li+solve+ (\fref{solver2}) more elegant.
+The type of computations \li+'a MWriter.t+ boils down (after functor
+application) to \li+(derivation list * state) option+. A computation in the monad
+represents a given point in the proof; the solver is focused on a rule; has
+proved a number of premises so far (the \li+derivation list+); has reached a
+certain \li+state+ (threaded through the premises). The \li+option+ type
+accounts for failure; in case a premise cannot be proved, the computation aborts
+and becomes \li+None+.
 
@@
 distinguish between an \li+outcome+ (the result of a call to \li+solve+) and a
-working state (a computation in the monad).
+working state (a computation in the \li+MWriter+ monad).
 
-An outcome is, as we mentioned earlier, the pair of a final \li+state+ along
+An \li+outcome+ is the pair of a final \li+state+ along
 with a \li+derivation+ that justifies that we reached this state. The pair is
-wrapped in \li+M.t+ (here, \li+option+): proving a formula may fail.
+wrapped in \li+M.t+ (here, \li+option+): if the computation of premises is a
+failure, then the proof of the desired goal is a failure too.
 
@@
 We now have a duality between the \li+outcome+ type (the result of solving a
-sub-goal) and the \li+m+ type (a computation within the monad, i.e. a working
+goal) and the \li+m+ type (a computation within the monad, i.e. a working
 state between two premises). Therefore, we introduce two high-level combinators:
@@
   \item \li+fail+ is for situations where no rule applies: this is a failed outcome.
+
+  \item \li+qed+ is a convenience combinator that pairs the state with the name
+    of the rule we want to conclude with; it makes the implementation of
+    \li+solve+ (\fref{solver2}) more elegant.
+
 \end{itemize}
@@
 \section{Backtracking}
+\label{sec:backtracking}
 
@@
 We thus augment \li+MOption+ with a search combinator (\fref{choice}), which in
-turn allows one to implement a high-level \li+choice+ combinator for our
+turn allows us to implement a high-level \li+choice+ combinator for our
 library. The \li+choice+ combinator attempts to prove a \li+goal+ by trying a
-function \li+f+ on several arguments \li+a+, each of which
-is associated to a given rule. We can add one more branch to the \li+solve+
-function, which attempts to prove a disjunction by first trying a
+function \li+f+ on several arguments of type \li+a+, each of which
+has a given \li+outcome+. We extend \li+solve+ with an extra case,
+which attempts to prove a disjunction by first trying a
 left-elimination (\Rule{Or-L}, \fref{proof-system2}), then a right-elimination
@@
 
-\section{Full example: quantifiers and derivations of logical rules}
+\section{Extension: quantifiers and proof trees}
 \label{sec:generate-logical}
 
-The final section is concerned with adding quantifiers to the language, and
-generating derivation trees that reflect the rules of the \emph{logic} rather
-than the rules of the \emph{algorithm}. The former is useful for debugging the
-inner workings of the algorithm; the latter is useful for producing a proof
-certificate.
+We mentioned earlier that the derivation we were building tracked the
+application of algorithmic rules; that is, we were building a \emph{trace} of
+the algorithm. While the trace is useful to extract information for the user,
+one may also want to build a proper proof witness in order to certify the
+validity of the formula.
+
+In order to make a proof tree relevant and not just provide the substitution as
+the proof witness, we introduce quantifiers to the language, and construct proof
+trees that apply the proof rules from \fref{logic}, \fref{proof-system2},
+\fref{logic-quantifiers}. The nodes of the proof tree are rules; each node is
+annotated, if applicable, by its implicit parameters. That is, \Rule{Refl} and
+\Rule{ExistsE} are annotated with their implicit $x$ parameter.
+
+\begin{figure}
+  \centering
+  \begin{mathpar}
+    \inferrule[ForallE]{
+      P
+    }{
+      \forall y.\ P
+    }
+  \end{mathpar}
+  \caption{Extra rule for the universal quantifier}
+  \label{fig:logic-quantifiers}
+\end{figure}
 
@@
 \begin{ocaml}
-(* Update of the [MyLogic] module. *)
+(* i) Update of the [MyLogic] module. *)
 type formula =
@@
 
-(* Update of the [Derivations] module. *)
+(* ii) Update of the [Derivations] module. *)
 type derivation =
@@
 
-(* Update of the [Combinators] module. *)
+(* iii) Update of the [Combinators] module. *)
 let prove (goal: Logic.formula) (x: ('a * rule_name) m): 'a outcome =
@@
 
-(* Update of the client. *)
+(* iv) Update of the client. *)
 let rec solve (state: state) (goal: formula): state outcome =
@@
 
-We augment the data type of formulas with quantifiers; we replace the type of
-rules with the rules from the logic. Furthermore, we demand that the \Rule{Refl}
-and \Rule{ExistsE} rules be annotated with their argument.
+The updates to the library required to implement quantifiers are minimal; the
+bulk of the work is essentially writing substitution and a proper treatment of
+binders on the client-side.
+
+\fref{quantifiers} presents in an informal style the series of updates
+required.
+\begin{enumerate}[i)]
+  \item We augment the data type of formulas with quantifiers; we replace the
+    type of rules with the rules from the logic. Furthermore, we demand that the
+    \Rule{Refl} and \Rule{ExistsE} rules be annotated with their argument.
 
     Bound variables are globally-unique atoms (strings); open variables are
-equivalence classes of the union-find, as before.
+    equivalence classes of the union-find, as before (not shown here).
 
-Now, a \li+goal+ is no longer just a formula: it is a formula interpreted in a
-given state. Indeed, since we now open binders and substitute bound variables
-for open variables, reading a formula only makes sense in a state where the open
-variables of the formula are properly bound in the environment.
+  \item We previously did not distinguish between a \li+goal+ and a
+    \li+formula+; this was only possible because we assumed all variables were
+    initially open, meaning that we could deference an open variable in any
+    \li+state+. Now, we open binders and substitute variables, through the
+    allocation of new points in the union-find (the \li+state+). Therefore, a
+    given \li+formula+ only makes sense when paired with a specific \li+state+.
 
-We thus change the \li+prove+ combinator to record the state upon creating
-a new node in the derivation tree. More precisely, the \li+prove+ combinator
-records the state after the premises have been satisfied. This is exactly what
-is needed to conclude a rule; for instance, in the \li+Exists+ case, the solver
-prods the union-find state to discover the instantiation choice that
-\emph{has been made} for the existentially-quantified variable, and records it
-in the proof tree.
+  \item We update the \li+prove+ combinator to record the \li+state+ upon
+    creating a new node in the derivation tree. More precisely, the \li+prove+
+    combinator records the \li+state+ after the premises have been satisfied.
 
-Here is a pretty-printed version of a derivation tree.
+  \item Only a slight is needed on the client side to record a proper proof
+    witness: in the \li+Exists+ case, the solver prods the union-find state to
+    discover the instantiation choice that \emph{has been made} for the
+    existentially-quantified variable, and records it in the proof tree.
+\end{enumerate}
+
+The sample code in the library comes with a pretty-printer. Here is the output
+for a simple formula that combines all features from our formula language.
 
@@
 
+\section{Source code}
+
+The library is available online at
+\url{https://github.com/msprotz/proof-search-monad}. The file \li+example01.ml+
+contains the full implementation of the primitive solver described in
+\sref{intro}. The file \li+example02.ml+ contains the backtracking solver
+written within the proof search monad, as described in \sref{backtracking}. One
+can get the non-backtracking version, described in \sref{derivations}, by
+replacing \li+MExplore+ with \li+MOption+. Finally, the file \li+example03.ml+
+contains the final algorithm described in \sref{generate-logical}. The
+representation of binders adopted in the last example is suboptimal; bound
+variables are represented using globally-unique atoms. One may want to use
+locally nameless, with De Bruijn indices for bound variables (as used in Mezzo).
+It allows keeps the boilerplate to a minimum, though.
+
+\section{Related work}
+
+ZOMG TODO
+
 \section{Conclusion}
 
-We presented a support library for writing a proof search engine using backtracking in
-any given logic; indeed, the library is parameterized by: the type of formulas; the type of
-rule applications; the internal state type of the client. By merely using the
+We presented a support library for writing a proof search engine using
+backtracking. The library is parameterized by: the type of formulas; the type of
+rule applications; the internal state type of the client. This leaves complete
+freedom for the client to define their own logic. By merely using the
 combinators of the library, the client gets derivations built for free; this allows

### 2015-10-13 paper/main.tex
As submitted.

@@
 \documentclass{easychair}
+
+\usepackage{fontspec}
+  \defaultfontfeatures{Mapping=tex-text}
+\usepackage{ucharclasses}
+  \setTransitionsFor{MathematicalOperators}{\begingroup\fontspec{DejaVu Sans}[Scale=MatchLowercase]}{\endgroup}
+\usepackage{xunicode}
+
+\usepackage{xltxtra}
 
@@
     escapeinside={(*}{*)},
-    morekeywords={above,abstract,adopts,alias,and,as,assert,begin,below,builtin,consumes,data,do,downto,duplicable,dynamic,else,empty,end,exclusive,explain,fact,fail,flex,for,from,fun,give,if,in,let,match,mutable,open,pack,perm,preserving,rec,take,taking,then,to,type,unknown,val,value,while,with,witness},
     deletekeywords={value},
@@
 derivation using rules from the logic. In our case, the rules are given in
-\fref{logic}; $[x/y]P$ means ``substitute $x$ with $y$ in $P$''. For
+\fref{logic}, where $[x/y]P$ means ``substitute $x$ with $y$ in $P$''. For
 instance, proving Equation \ref{ex1} requires applying \Rule{ExistsE}, then
@@
 The library has been used, in a preliminary form, to implement the core of the
-Mezzo type-checker. This paper presents a cleaned-up, isolated version of this
-library that exposes a proper interface using monads and domain-specific
-combinators.
+Mezzo type-checker~\cite{protzenko-phd-14}. This paper presents a cleaned-up,
+isolated version of this library that exposes a proper interface using monads
+and domain-specific combinators.
 
@@
 The logic we present is a much simplified version of the logic (type system) of
-Mezzo. In this document, we only mention the right-exists quantifier. General
-systems such as Mezzo have all four possible combinations of left/right
-exists/forall.
+Mezzo~\cite{pottier-protzenko-13}. In the present document, we only mention the
+right-exists quantifier. General systems such as Mezzo have all four possible
+combinations of left/right exists/forall.
 %
@@
 goal is $\forall x, \exists \f y, \forall z.\ P$, picking $\f y = z$ makes no
-sense. Mezzo forbids this choice using \emph{levels}; in the present document,
+sense. Mezzo forbids this choice using \emph{levels}~\cite{pottier-remy-emlti}; in the present document,
 we skip this discussion altogether and assume that ``all is well''. Finally,
@@
 
-We implement proof search in OCaml (\fref{formulas}); we implement substitutions
-using a union-find data structure. The data type of formulas
+We implement proof search in OCaml~\cite{ocaml} (\fref{formulas}); we implement substitutions
+using a union-find data structure~\cite{cormen-en,tarjan-75}. The data type of formulas
 is self-explanatory. Variables are implemented as equivalence classes in the
@@
 
-\fref{solver} implements a solver for our minimal problem; written within the
-\li+MOption+ monad, it returns either \li+Some state+ (in case a successful
-substitution has been found), or \li+None+ if no solution exists. The solver is
+\fref{solver} implements a solver for our minimal problem; since we perform
+computations that either return a result of a failure, the code leads itself
+well to an implementation using
+monads~\cite{wadler-comprehending-92,wadler-essence-92}, in our case, the
+\li+MOption+ monad. The \li+Some state+ is for success, meaning a substitution
+has been found, while the \li+None+ case means no solution exists. The solver is
 complete.
@@
 proper derivation. The former is still faithfully implemented by the option
-monad; the latter is implemented by the writer monad.
+monad; the latter is implemented by the writer monad~\cite{jones-95-}.
 
@@
 
-ZOMG TODO
+An article titled ``The Proof Monad'' already exists~\cite{kirchner-munoz-10};
+in spite of the closely related title, the article is concerned with a slightly
+different problem, namely giving an operational semantics to tactic languages
+used in theorem provers. In that sense, the article is related to
+Mtac~\cite{ziliani-13}, which is also concerned with a proper monad for writing
+tactics in Coq.
+
+Hedges~\cite{hedges-14} compares various explorations monads, notably using the
+continuation monad, the selection monad~\cite{escardo-2010} and their respective
+monad transformers. The main focus of the article seems to be the relationship
+between backtracking and game theory.
+
+Hinze~\cite{hinze-00} shows how to use the backtracking monad transformer, i.e.
+add backtracking to any existing monad. It would be interesting to determine
+whether our library can be re-implemented using a backtracking monad
+transformer, rather than the writer transformer applied to the monad of
+non-determinism. The (draft) version of the library used in Mezzo also builds
+failed derivations (as error messages) that list all attempted proofs, along
+with the first premise that failed; doing so would not be possible using
+exceptions.
+
+The \li+choice+ operator is related to polarization and
+focusing~\cite{liang-miller-07}. For instance, in the problem $\f y = x \vee \f
+y = z$, depending on which side of the disjunction the algorithm considers first,
+the outcome is going to be different. This is analogous to a synchronous phase
+(where the order of the rules matters, and where a particular choice may have
+consequences on the rest of the search). Similarly, one may swap premises
+chained by the \li+>>=+ operator, as the order doesn't matter. This is analogous
+to a asynchronous phase.
 
@@
 
-\bibliographystyle{abbrvnat}
-\bibliography{local}
+\bibliographystyle{plain}
+\bibliography{english,local}
 

### 2015-11-10 paper/main.tex
First pass on the reviewers' comments.

@@
   \let\TirName\textsc
-  \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
+  % \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
   \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
-  \let\DefRule\RefTirName
+  % \let\DefRule\RefTirName
   \let\Rule\DefTirName
@@
 \label{sec:intro}
+
+This paper attempts to present, in a tutorial-style, the design of an OCaml
+library. In order to facilitate the discussion, we focus on a very constrained
+logic; later (\sref{extending}), we mention how to extend the library to
+cover more use-cases. The original motivation for the library was to serve as a
+core building block for the type-checker of Mezzo~\cite{protzenko-phd-14}. The
+nature of the core, minimal logic that we are about to present is, of course,
+inspired by typical nature of type-checking problems: it features equality,
+quantifiers, and positive literals; \sref{extending} mentions how to extend it
+with function symbols and variance (negative positions), as is typical for
+type-checking problems.
 
@@
       prove goal begin
+        (* Recursive call reduces to the [axiom ... R_Refl] case below. *)
         premise (prove_equality state goal v1 v2) >>=
@@
 
+\section{Extending the library; extending the logic; limitations}
+\label{sec:extending}
+
+The approach advocated in the present paper works well for sequent-style
+calculi; in the context of Mezzo, the logic is extended with the following extra
+features:
+\begin{itemize}
+  \item function symbols, such as, but not limited to: ML arrow types ($\to$),
+    type applications ($\mathsf{list}\ \alpha$), constructor applications
+    ($\mathsf{Cons}\;\{ \mathsf{head}:\alpha;\ \mathsf{tail}:\mathsf{list}\
+    \alpha\}$)
+  \item positive and negative positions (also known as ``variance'' in
+    type-checking lingo; this applies to function symbols, such as arrows or
+    type constructors)
+  \item higher-order quantification ($\forall (p: \mathsf{predicate}) \ldots$)
+  \item affinity (where some hypothesis may be used at most once)
+  \item framing, which bears some similarities with focusing.
+\end{itemize}
+
+This in turn requires the client code to keep track of more information, while
+also adopting more sophisticated data structures. In particular, the client code
+now carries, in addition to a substitution (\emph{a.k.a.} union-find), a
+\emph{set} of available hypotheses, which flows left-to-right in the algorithmic
+rules. Changing polarities changes the direction of the flow.
+
+For the complete system (Mezzo), there are many more backtracking points. The
+client code choses to backtrack on only a subset of them.
+
+A limitation of this library is that it only works as long as every branch of
+the exploration terminates; contrary to Kiselyov \emph{et al}.'s
+library~\cite{oleg05}, we do not implement fair interleaving. One could
+conceivably bound the depth of the search tree, but the exploration of the tree
+remains sequential, not concurrent.
+
+If one is willing to give up on modularity, stronger static guarantees can be
+attained by making the \li+rule_name+ type more specific; namely, by encoding in
+each constructor the number of premises required. The drawback is that the
+library now has to be aware of the specific logic; in the current state of
+things, the library is completely agnostic with regards to the client code's
+particular logical system.
+
+It is unclear how one could memoize the sub-computations; the state of the
+client code (the union-find) may not change very much in the simplified logic
+presented here; however, for a more evolved logic, it is likely that the client
+state evolves monotonically, and makes memoizing difficult, as the result of
+sub-calls depends very much on the state.
+
 \section{Source code}
@@
 
+\section*{Acknowledgments}
+
+I wish to thank Gabriel Scherer who helped me sketch the initial version of the
+library; François Pottier for motivating me enough that I would want to write
+about it, as well as providing the persistent union-find implementation; and
+Anonymous Review \#2 for a very high quality review with numerous excellent
+suggestions.
+
 \bibliographystyle{plain}

### 2015-11-10 paper/main.tex
More excellent remarks from reviewer #2.

@@
   \let\TirName\textsc
-  % \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
+  \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
   \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
-  % \let\DefRule\RefTirName
+  \let\DefRule\RefTirName
   \let\Rule\DefTirName
@@
 library. In order to facilitate the discussion, we focus on a very constrained
-logic; later (\sref{extending}), we mention how to extend the library to
+logic; later (\sref{extending}), we briefly discuss how to extend the library to
 cover more use-cases. The original motivation for the library was to serve as a
@@
 quantifiers, and positive literals; \sref{extending} mentions how to extend it
-with function symbols and variance (negative positions), as is typical for
-type-checking problems.
+with, among other things, function symbols and variance (positive/negative
+positions), as is typical for type-checking problems.
 
@@
 equalities, one may want to use the rules from \fref{proof-system}. These rules
-differ from \fref{logic} in that they are algorithmic; they take an input and
-return an output.
+differ from \fref{logic} in that they are algorithmic; they take an input
+($\vdash$) and
+return an output ($\dashv$).
 
@@
     }{
-      V \vdash P \dashv \sigma'
+      V, \sigma \vdash P \dashv \sigma'
     }
@@
     }{
-      V \vdash P \wedge Q \dashv \sigma''
+      V, \sigma \vdash P \wedge Q \dashv \sigma''
     }
@@
 
-let unify state v1 v2: state =                  let rec solve state formula: state =
+let unify state v1 v2: state option =           let rec solve state formula: state =
   match P.find v1 state, P.find v2 state with     match formula with
@@
 
-The type of computations \li+'a MWriter.t+ boils down (after functor
-application) to \li+(derivation list * state) option+. A computation in the monad
+The type of computations \li+'a MWriter.t+ boils down to \li+(derivation list * state) option+
+after functor application. A computation in the monad
 represents a given point in the proof; the solver is focused on a rule; has
@@
   \item function symbols, such as, but not limited to: ML arrow types ($\to$),
-    type applications ($\mathsf{list}\ \alpha$), constructor applications
+    type applications ($\mathsf{list}\ \alpha$) and constructor applications
     ($\mathsf{Cons}\;\{ \mathsf{head}:\alpha;\ \mathsf{tail}:\mathsf{list}\
@@
 
-For the complete system (Mezzo), there are many more backtracking points. The
-client code choses to backtrack on only a subset of them.
-
 A limitation of this library is that it only works as long as every branch of
@@
 remains sequential, not concurrent.
+
+The logic does not necessarily have to be decidable for this approach to work
+well; indeed, we conjecture that type-checking Mezzo programs is not
+decidable~\cite[p.  167]{protzenko-phd-14}. What matters is that exploration
+follows a deterministic set of rules; in Mezzo, the backtracking points are
+chosen and controlled~\cite[p. 165]{protzenko-phd-14}. We, of course, explore a
+fine-tuned subset of the search space.
 
@@
 
-It is unclear how one could memoize the sub-computations; the state of the
-client code (the union-find) may not change very much in the simplified logic
-presented here; however, for a more evolved logic, it is likely that the client
-state evolves monotonically, and makes memoizing difficult, as the result of
-sub-calls depends very much on the state.
+It is unclear how one could memoize the sub-computations, as they depend very
+much on the current state, which is likely to change at every step.
 

### 2015-11-27 pres/pres.tex
Presentation.

\documentclass[final,xetex]{beamer}

\input{talks/common/macros.tex}
\input{papers/common/macros.tex}
\input{bubbles.tex}

% ------------------------------------------------------------------------------

\usepackage{ucharclasses}
  \setTransitionsFor{Dingbats}{\begingroup\fontspec{DejaVu Sans}[Scale=MatchLowercase]}{\endgroup}
\usepackage{multirow}
\usepackage{mathtools}
\usepackage[normalem]{ulem}

%\def\mintedusecache{}
\def\mintedwritecache{}

% ------------------------------------------------------------------------------

\newcommand{\sidenote}[1]{
\raggedleft
\footnotesize
\itshape
#1
}

\newcommand{\statement}[1]{
  \begin{frame}[plain]
  \begin{center}
    \Large
    \textcolor{skyblue3}{
    #1
    }
  \end{center}
  \end{frame}
}

\newcommand{\V}{\ensuremath{\mathcal{V}}} % The prefix
\newcommand{\R}{\ensuremath{\mathcal{R}}} % Rigid
\newcommand{\F}{\ensuremath{\mathcal{F}}} % Flexible

% ------------------------------------------------------------------------------

\TangoColors
\mode<presentation> {
  \setbeamertemplate{navigation symbols}{}
  \setbeamertemplate{items}[circle]
  \setbeamercolor*{Title bar}{bg=white,fg=chameleon3}
  \setbeamersize{text margin left=2em}
  \setbeamertemplate{footline}{
    \leavevmode%
    \hbox{%
    \begin{beamercolorbox}[wd=.3\paperwidth,ht=2.25ex,dp=1ex,center]{author in head/foot}%
      \usebeamerfont{author in head/foot}\insertshortauthor\ —\ \insertshortinstitute
    \end{beamercolorbox}%
    \begin{beamercolorbox}[wd=.4\paperwidth,ht=2.25ex,dp=1ex,center]{title in head/foot}%
      \usebeamerfont{title in head/foot}\insertshorttitle
    \end{beamercolorbox}%
    \begin{beamercolorbox}[wd=.3\paperwidth,ht=2.25ex,dp=1ex,right]{date in head/foot}%
      \usebeamerfont{date in head/foot}\insertshortdate{}\hspace*{2em}
      \insertframenumber{} / \inserttotalframenumber\hspace*{2ex}
    \end{beamercolorbox}}%
    \vskip0pt%
  }
}

\title[The Proof Search Monad]{
  Functional Pearl: the Proof Search Monad
}
\author[J. Protzenko]{
  \newauthor{Jonathan Protzenko}{protz@microsoft.com}
}

\institute{Microsoft Research}
\date{Nov. 23th, 2015}

\begin{document}

\begin{frame}
  \begin{center}
    {\raggedright\large (functional pearl)\\
    \Huge\vspace{.2em}
    \hspace{1.2em}\green{\emph{the Proof Search Monad}}}
  \end{center}

  \vspace{3cm}

    Jonathan Protzenko (Microsoft Research)\\
    \scriptsize
    \texttt{protz@microsoft.com}
\end{frame}

\begin{frame}{The Proof Search Monad}
  A \green{library} developed to implement the type-checker of \mezzo.

  \bigskip

  In this talk:
  \begin{itemize}
    \item what is \mezzo (from a very high-level)
    \item generalize (when is this library suitable)
    \item the library itself (combinators and implementation).
  \end{itemize}
\end{frame}

\begin{frame}[plain]
  \backupslide
  \MzTitle{Where it all came from}
\end{frame}

\begin{frame}{My thesis in one slide}
  ``Separation logic as a type system''. % who's familiar with it?

  \begin{itemize}
    \item \green{\mezzo}: \emph{barely} a type system
      {\tiny(flow-sensitive, structural information, keeps track of local
      aliasing)}
    \item Hindley-Milner: \green{nope}
    \item Because: \green{undecidable} {\tiny(System F, entailment, framing,
      higher-order logic)}
    \item But: we \emph{still} want \green{inference}
    \item So: \emph{\green{heuristics}} (it ``mostly'' works)
  \end{itemize}
\end{frame}

\begin{frame}[fragile]{Flow-sensitivity}

  \mezzo has \green{permissions}, of the form $\tyatomic xt$, separated by
  $\ast$.

  \bigskip

  \begin{description}[\hspace{10em}]
    \item[In ML:] $\Gamma = x: t, y: u$
    \item[In \mezzo:] $P = \tyatomic xt \ast \tyatomic yu$
  \end{description}

  \bigskip

  \bigskip

  \MzStartBubbles
\begin{mzcode}
val f (x: ...): ... = #\MzProgramPoint#
  let y = ... in #\MzProgramPoint#
  ...#\MzProgramPoint#
\end{mzcode}

  \MzStartBubbles
  \MzAutoBubble{2em}{2em}{1cm}{}{%
    P₁
  }
  \MzAutoBubble{0em}{3em}{1cm}{}{%
    P₂
  }
  \MzAutoBubble{2em}{-2em}{1cm}{}{%
    P₃
  }

  \bigskip

  \bigskip

  \only<5>{
  This allows keeping track of \green{ownership}.
  }

\end{frame}

\begin{frame}[fragile]{Why ownership?}
  At each program point, the programmer knows which objects they own, and how
  they own them.
  Ownership is either shared (duplicable) or unique (exclusive).

  \vfill

  \begin{description}
    \item[Data-race freedom]
      \emph{Mutable} objects have a unique owner. Therefore, at most one thread
      may mutate an object at any given time. \mezzo programs are data-race
      free.
    \item[State change]
      If I'm the sole owner of an object, I'm allowed to break its invariants
      (type). Important because of fine-grained aliasing tracking.
  \end{description}
\end{frame}

\begin{frame}[fragile]{Ownership example}

  Data-race \green{freedom} \emph{and} \green{state change}.

  \bigskip

  \bigskip

  \begin{mzcode}
let r = newref x in
(* r @ Ref { contents = x } *)
r := y;
(* r @ Ref { contents = y } *)
  \end{mzcode}

\end{frame}

\begin{frame}{A rich type system\ldots}
  \begin{description}[\hspace{10.2em}]
    \item[Singleton types] \code{x @ (=y)}: \code{x} \green{is} \code{y}
      \\ Written as: $x = y$
    \item[Constructor types] \code{xs @ Cons \{ head: t; tail: u \}}
      (special-case: \code{t} is a singleton, we write
      \code{xs @ Cons \{ head = …; tail = … \}})
    \item[Decomposition] via \green{unfolding} (named fields),
      \green{refinement} (matching) and \green{folding} (subtyping)
    \item[Several possible types] \code{x @ (int, int)},\\
      \code{x @ ∃(y,z: value).\\~~(=y | y @ int, =z | z @ int)},
      \\\code{x @ ∃t.t}, etc.
  \end{description}
\end{frame}

\begin{frame}{\ldots that still isn't quite a logic}
  \mezzo remains a type system.
  \begin{itemize}
    \item far less connectives and rules
    \item $\tyatomic f{\tyarrow tu} \ast \tyatomic xt \not\leq
      \tyexists y\kterm {\tyatomic yu}$ (no implicitly callable ghost functions)
    \item no built-in disjunction (only tagged sums)
  \end{itemize}

  \bigskip

  \mezzo's type system feels like a limited fragment of \green{intuitionistic
  logic}.
\end{frame}

\begin{frame}{Subtraction: an unusual algorithm}
  \begin{itemize}
    \item Subtyping needs to be decided for \green{function calls} and for
      \green{function bodies}.
    \item %Subsumption rules allow instantiating quantifiers \green{implicitly}:
      Blurs the frontier between type-checking and logics.
    \item %Hence,
      The subtyping algorithm \emph{has to} \green{perform
      inference}%: introducing existentials in the goal, introducing universals
      %in the hypothesis (current permission).
  \end{itemize}
\end{frame}

\begin{frame}{More about subtraction}
  The operation is written $P - Q = R$ and assumes $P$ has been
  \emph{normalized} (all left-invertible rules have been applied).

  \bigskip
  \bigskip
  \bigskip

  {\Large
  \[
    \tikz[remember picture, baseline=-0.65ex] \node(sub-v) {$\mathcal{V}$};
    (
    \tikz[remember picture, baseline=-0.65ex] \node(sub-p) {P};
    -
    \tikz[remember picture, baseline=-0.65ex] \node(sub-q) {Q};
    ) =
    \tikz[remember picture, baseline=-0.65ex] \node(sub-vp) {$\mathcal{V'}$};
    \tikz[remember picture, baseline=-0.65ex] \node(sub-r) {R};
  \]
  }
  \begin{tikzpicture}[
    remember picture,
    overlay
  ]
    % V
    \node[
      draw, ultra thick, aluminium5, ellipse, fill=aluminium1
    ] (label-v) at ($(sub-v)+(-3em,-2em)$) {Binders};
    \draw[
      ultra thick, aluminium5, ->
    ] (label-v.north) to [out=90,in=180] (sub-v.west);
    \draw[
      ultra thick, aluminium5, ->
    ] (label-v.east) to [out=0,in=270] (sub-vp.south);

    % P
    \node[
      draw, ultra thick, aluminium5, ellipse, fill=aluminium1
    ] (label-p) at ($(sub-p)+(-3em,3em)$) {Hypothesis};
    \draw[
      ultra thick, aluminium5, ->
    ] (label-p.south) to [out=270,in=90] (sub-p.north);

    % Q
    \node[
      draw, ultra thick, aluminium5, ellipse, fill=aluminium1
    ] (label-q) at ($(sub-q)+(3em,3em)$) {Goal};
    \draw[
      ultra thick, aluminium5, ->
    ] (label-q.south) to [out=270,in=90] (sub-q.north);

    % R
    \node[
      draw, ultra thick, aluminium5, ellipse, fill=aluminium1
    ] (label-r) at ($(sub-r)+(3em,-2em)$) {Remainder};
    \draw[
      ultra thick, aluminium5, ->
    ] (label-r.north) to [out=90,in=0] (sub-r.east);

  \end{tikzpicture}

  \bigskip
  \bigskip

  This means:\\
  \qquad``with the instantiation choices from $\V'$, we get $P \leq Q \ast R$''.

\end{frame}

\begin{frame}{Subtraction example (1)}
  In order to type-check the application $f\ x$, we ask:
  \[
    P - \tyatomic f{\tyarrow {\alpha}{\beta}} \ast
        \tyatomic x{\alpha}\quad=\quad?
  \]

  \bigskip

  The algorithm must guess $\alpha$, $\beta$, and the remainder.
\end{frame}

\begin{frame}{Subtraction example (2)}
  \[
  \begin{array}{ll}
  & \tyatomic \ell{\tyconcrete \kcons{ \khead = h; \ktail = t }} \ast \\
  &\tyatomic h{\tyref\tyint} \ast \tyatomic t{\tylist {(\tyref\tyint)}} \\
  - \quad \\
  &\tyatomic \ell{\tylist{(\tyref\tyint)}} \\
  =  \quad\\
  &\tyatomic \ell{\tyconcrete \kcons{ \khead = h; \ktail = t }}
\end{array}
  \]
\end{frame}

\begin{frame}{Subtraction example (3)}
 \[
  \begin{array}{ll}
  & \tyatomic z{\exists\alpha, \forall\beta. (\alpha, \beta)} \\
  - \quad \\
  & \tyatomic z{\exists\alpha', \forall\beta'. (\alpha', \beta')} \\
  = \quad \\
  & ?
  \end{array}
\]
\end{frame}

\begin{frame}{Backtracking}
  Inference uses \emph{flexible} variables.

  \bigskip

  There may be \green{several solutions}:
  \[
  \tyatomic x\tyint - \tyatomic x\alpha =
  \tyatomic x\tyint\quad\text{ with }\quad
  \begin{cases}
    \alpha = \tyint \\
    \alpha = \tysingleton x \\
    \alpha = \top
  \end{cases}
  \]
  Not all solutions are explored: $\alpha$ could be $\tybar\beta p$.

  \bigskip

  There are many other backtracking points: right-introduction \emph{vs.}
  left-introduction, which atomic permission to focus\ldots
\end{frame}

\begin{frame}[fragile]{How to implement all that?}
  A module that takes care of running the proof search and providing an answer
  to:
  \[
  P-Q\quad = \quad?
  \]

  \bigskip

  \begin{minted}[fontsize=\footnotesize]{ocaml}
| TyQ (Forall, binding1, _, t'1), TyQ (Exists, binding2, _, t'2) ->
    par env judgement "Intro-Flex" [
      try_proof_root "Forall-L" begin
        let env, t'1, _ = bind_flexible_in_type env binding1 t'1 in
        sub_type env t'1 t2 >>=
        qed
      end;
      try_proof_root "Exists-R" begin
        let env, t'2, _ = bind_flexible_in_type env binding2 t'2 in
        sub_type env t1 t'2 >>=
        qed
      end
    ]
  \end{minted}
\end{frame}

\begin{frame}[plain]
  \backupslide
  \MzTitle{The proof search monad}
\end{frame}

\begin{frame}{}
  \begin{itemize}
    \item \mezzo is \green{not decidable}; however
    \item we \emph{know} \green{where} we want to backtrack;
      {\tiny(and where we do not want to go, e.g. $\pi \ast \pi' \ast \pi'' \leq p \ast q \ast r$)}
    \item we \emph{know} \green{which} branch is most likely to succeed;
      {\tiny($\alpha = \tyint$ usually better than $\alpha = \top$)}
    \item we \emph{know} that sub-branches \emph{terminate}
      {\tiny(no need to interleave)}
  \end{itemize}
\end{frame}

\begin{frame}{}
  In that case, adopt the library style where:
  \begin{itemize}
    \item the proof search algorithm looks \green{like the paper rules};
    \item you get a \green{derivation for free};
    \item the library works for \green{any logic} {\tiny(not in \mezzo, hence this
      paper)}
  \end{itemize}
\end{frame}

\begin{frame}[plain]
  \backupslide
  \MzTitle{The library, dissected}
\end{frame}

\begin{frame}[fragile]{From the library's perspective (1)}
  The client's logic must satisfy \code{LOGIC}.

  \bigskip

  \begin{minted}{ocaml}
module type LOGIC = sig
  type formula
  type rule_name
  type state
end
  \end{minted}
\end{frame}

\begin{frame}[fragile]{From the library's perspective (2)}
  The library defines derivation trees for \code{LOGIC}.

  \bigskip

  \begin{minted}{ocaml}
module Derivations = functor (L: LOGIC) -> struct
  type derivation = goal * rule
  and goal = L.state * L.formula
  and rule = L.rule_name * premises
  and premises = Premises of derivation list
end
  \end{minted}
\end{frame}

\begin{frame}[fragile]{From the library's perspective (3)}
  An \code{'a m} is the working \green{state} of a rule application.

  \vfill

  \begin{minted}[fontsize=\footnotesize]{ocaml}
module Make(Logic: LOGIC)(M: MONAD) = struct
  module Proofs = Derivations(Logic)
  include Proofs

  module L: MONOID with type t = Proofs.derivation list = struct
    type t = Proofs.derivation list
    let empty = []
    let append = List.append
  end

  include WriterT(M)(L) (*
    type 'a m = (L.t * 'a) M.m
    val return : 'a -> 'a m
    val tell : L.a -> unit m
    val ( >>= ) : 'a m -> ('a -> 'b m) -> 'b m
  *)

  ...
  \end{minted}
\end{frame}

\begin{frame}[fragile]{From the library's perspective (4)}
  An \code{'a outcome} is the \green{result} of a rule application.

  \bigskip

  \begin{minted}[fontsize=\footnotesize]{ocaml}
  ...

  type 'a outcome = ('a * derivation) M.m

  (* _Record_ a proof in the premises. *)
  val premise: 'a outcome -> 'a m
  (* _Conclude_ from the given premises. *)
  val prove:
    Logic.formula ->
    (Logic.state * rule_name) m ->
    Logic.state outcome
end
  \end{minted}
\end{frame}

\begin{frame}[fragile]{From the client's perspective (1)}
  \begin{minted}[fontsize=\footnotesize]{ocaml}

module MyLogic = struct ... end
module MyMonad = ProofSearchMonad.Make(MyLogic)(MExplore)

let rec solve (state: state) (goal: formula): state outcome =
  match goal with
  ...
  | And (g1, g2) ->
      prove goal begin
        premise (solve state g1) >>= fun state ->
        premise (solve state g2) >>= fun state ->
        state, R_And
      end
  ...
  \end{minted}
\end{frame}

\begin{frame}[fragile]{Let's refine (1)}
  \begin{minted}[fontsize=\footnotesize]{ocaml}
let qed rule = fun state -> state, rule

let rec solve (state: state) (goal: formula): state outcome =
  match goal with
  ...
  | And (g1, g2) ->
      prove goal begin
        premise (solve state g1) >>= fun state ->
        premise (solve state g2) >>=
        qed R_And
      end
  ...
  \end{minted}
\end{frame}

\begin{frame}[fragile]{Let's refine (2)}
  \begin{minted}[fontsize=\footnotesize]{ocaml}
val choice :
  Logic.formula ->
  'a list -> ('a -> (Logic.state * rule_name) m) -> Logic.state outcome

let rec solve (state: state) (goal: formula): state outcome =
  match goal with
  ...
  | Or (g1, g2) ->
      choice goal [ R_OrL, g1; R_OrR, g2 ] (fun (rule, g) ->
        premise (solve state g) >>=
        qed rule
      )
  ...
  \end{minted}
\end{frame}

\begin{frame}{In conclusion}
  \begin{itemize}
    \item A powerful library for writing the proof search of \mezzo
    \item Generalizes if your problem fits the earlier description
    \item Proof derivations for free
    \item We used them mostly for debugging {\tiny(In \mezzo, a post-processing
      phase tries to extract \emph{relevant} parts of a \emph{failed}
      derivation.)}
  \end{itemize}
\end{frame}

\begin{frame}{Next}
  \begin{itemize}
    \item \green{Extend} the library to record \emph{failed} derivations
      {\tiny(\code{choice} now records \emph{all} the things we tried; \code{>>=}
      records up to the first \emph{failed} premise)}
    \item See if compatible with more complex exploration strategies.
  \end{itemize}
\end{frame}

\end{document}

\begin{frame}{A glance at the type-checking rules}
  General form: $K, P \vdash e: t$. (K = kinding environment)

  \bigskip

  \small
  \begin{mathpar}
    \inferrule[Sub]
    {K; P_2 \vdash e: t_1 \\\\ P_1 \leq P_2 \\ t_1 \leq t_2}
    {K; P_1 \vdash e: t_2}

    \inferrule[Frame]
    {K; P_1 \vdash e: t}
    {K; P_1\ast P_2 \vdash e: \tybar{t}{P_2}}

    \inferrule[Read]
    {t \text{ is duplicable} \\\\
     P \text{ is } \tyatomic{x}{
                     \tyconcrete{A}{\ldots; f: t; \ldots}}
    }
    {K; P \vdash x.f: \tybar{t}{P}}

    \inferrule[Tuple]
    {}
    {K \vdash (x_1, \ldots, x_n): (\tysingleton {x_1}, \ldots, \tysingleton{x_n})}

    \inferrule[Application]
    {}
    {K; \tyatomic{x_1}{t_2 \rightarrow t_1} \ast \tyatomic{x_2}{t_2} \vdash \eapply{x_1}{x_2}: t_1}
  \end{mathpar}
\end{frame}

\begin{frame}{A glance at the subsumption relation}
  \small
  \begin{mathpar}
  \inferrule[DecomposeTuple]
    {}
    {
    \begin{array}{r@{}l}
    & \tyatomic{y}{(\ldots,t,\ldots)} \cr
    \subsub
    \tyexists x\kterm{(
      \tystar
        {& \tyatomic{y}{(\ldots, \tysingleton{x}, \ldots)}}
        {\tyatomic{x}{t}}
      )}
    \end{array}
    }

  \inferrule[EqualsForEquals]
    {}
    {
      \tystar
      {(y_1 = y_2)}
      {[y_1/x]P}
      \subsub
      \tystar
      {(y_1 = y_2)}
      {[y_2/x]P}
    }

  \inferrule[EqualityReflexive]
    {} % si x : term
    {\tyempty \leq (x = x)}

  \inferrule[Fold]
    {\tyconcrete{A}{\vec{f}: \vec{t}} \text{ is an unfolding of } \tyapp{X}{\vec{T}}}
    {
      \tyatomic{x}{\tyconcrete{A}{\vec{f}: \vec{t}}}
      \leq
      \tyatomic{x}{\tyapp{X}{\vec{T}}}
    }
  \end{mathpar}
\end{frame}

\begin{frame}{A glance at the subsumption relation (2)}
  \small
  \begin{mathpar}
  \inferrule[ForallElim]
    {}
    {\tyforall X\kappa P \leq [T/X]P}

  \inferrule[CopyDup]
    {P\text{ is duplicable}}
    {C[t] \ast P \leq C[\tybar tP] \ast P}

  \inferrule[HideDuplicablePrecondition]
    {P \text{ is duplicable}}
    {
      (\tyatomic{x}{\tyarrow{\tybar{t_1}{P}}{t_2}})
      \ast P
      \leq
      \tyatomic{x}{\tyarrow{t_1}{t_2}}
    }

  \inferrule[ExistsIntro]
    {}
    {[T/X]P \leq \tyexists X\kappa P}

  \inferrule[CoArrow]
    {u_1 \leq t_1 \\ t_2 \leq u_2}
    {\tyatomic{x}{\tyarrow{t_1}{t_2}} \leq \tyatomic{x}{\tyarrow{u_1}{u_2}}}

  \inferrule[Unfold]
    {\tyconcrete{A}{\vec{f}: \vec{t}} \text{ is an unfolding of } \tyapp{X}{\vec{T}} \\\\
     \tyapp{X}{\vec{T}} \text{ has only one branch}}
    {
      \tyatomic{x}{\tyapp{X}{\vec{T}}}
      \leq
      \tyatomic{x}{\tyconcrete{A}{\vec{f}: \vec{t}}}
    }
  \end{mathpar}
\end{frame}

\begin{frame}{A suitable representation of permissions}
  \mezzo is a powerful language: the type-checker is complex, because of the
  interaction between:
  \begin{itemize}
    \item \green{duplicable} \emph{vs.} \green{non-duplicable} permissions,
    \item equivalent permissions: \\
      \code{\green{z @ (=x, =y)} * x @ ref int * y @ ref int ≡ \\
      z @ (ref int, ref int)},
    \item inference (of type application): \code{cons [?] (x, y)},
    \item subtyping: \\
      \code{[a] duplicable a => (ref a) -> a ≡ \\
      \empty[y: value] (ref (=y)) -> (=y)},
    \item the frame rule…
  \end{itemize}
\end{frame}

\begin{frame}{The prototype}
  Backtracking \green{stops} at the expression level: we keep \green{one
  solution} when type-checking an expression.

  \bigskip

  The implementation relies on:
  \begin{itemize}
    \item efficient (good complexity) and easy-to-use (persistent) \green{data
      structures} for inference with backtracking (union-find, levels)
    \item fine-tuned \green{heuristics} (prioritize more likely solutions first)
  \end{itemize}
  Both required significant effort.

%   \bigskip

%   \sidenote{(Some performance limitations because of
%   binders.)}
\end{frame}

\begin{frame}[fragile]{\mezzo: reason about ownership}
  Consider list concatenation.

  \bigskip

  \begin{mzcode}
val append: [a] (
  consumes xs: list a,
  consumes ys: list a
) -> (zs: list a)
  \end{mzcode}
\end{frame}

\begin{frame}[fragile]{\mezzo: reason about ownership}
  \green{Example}: \code{list (ref int)}.

  \bigskip

  \bigskip

  \MzStartBubbles
  \begin{mzcode}
...#\MzProgramPoint#
let zs = append (xs, ys) in
...#\MzProgramPoint#
  \end{mzcode}

  \MzStartBubbles
  \MzAutoBubble{6em}{3em}{4cm}{Before function call}{%
    xs @ list (ref int) * ys @ list (ref int)
  }
  \MzAutoBubble{6em}{-3em}{4cm}{After function call}{%
    zs @ list (ref int)
  }
\end{frame}

\begin{frame}[fragile]{\mezzo: reason about ownership}
  \green{Example}: \code{list int}.

  \bigskip

  \MzStartBubbles
  \begin{mzcode}
...#\MzProgramPoint#
let zs = append (xs, ys) in
...#\MzProgramPoint#
  \end{mzcode}
  \MzStartBubbles
  \MzAutoBubble{6em}{3em}{6cm}{Before function call}{%
    xs @ list int * ys @ list int
  }
  \MzAutoBubble{6em}{-3em}{6cm}{After function call}{%
    xs @ list int * ys @ list int *\\
    zs @ list int
  }

\end{frame}

\end{document}
